---
title: "Adding Variables To Data"
author: "Belle Schmidt"
date: "2024-12-31"
output: html_document
---

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
library(readr)
```


```{r}
# reading in necessary data
plays_of_WR <- read.csv("wr_plays.csv")
predicting_plays <- read.csv("offense_prediction_plays.csv")
games <- read.csv("games.csv")
movement_data <- read.csv("movement_data.csv")
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_8.csv")
tracking_week9 <- read.csv("tracking_week_9.csv")
```

# Working With Tracking Data ---------------------------------------

## Calculating Starting Position, Snap Position, and Distance Traveled From Starting Position to Snap Position For All Players
```{r}
# week 1 tracking data
movement_week1 <- tracking_week1 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 2 tracking data 
movement_week2 <- tracking_week2 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 3 tracking data 
movement_week3 <- tracking_week3 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 4 tracking data 
movement_week4 <- tracking_week4 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 5 tracking data 
movement_week5 <- tracking_week5 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 6 tracking data 
movement_week6 <- tracking_week6 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 7 tracking data 
movement_week7 <- tracking_week7 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 8 tracking data 
movement_week8 <- tracking_week8 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

 # week 9 tracking data 
movement_week9 <- tracking_week9 |> 
  filter(frameId == 1 | frameType == "SNAP") |> 
  select(gameId:displayName, frameType, x, y) |> 
  pivot_wider(names_from = frameType, values_from = x:y) |> 
  rename(starting_position_x = x_BEFORE_SNAP,
         starting_position_y = y_BEFORE_SNAP,
         snap_position_x = x_SNAP,
         snap_position_y = y_SNAP) |> 
  mutate(distance_moved = sqrt((snap_position_x - starting_position_x) ** 2 + (snap_position_y - starting_position_y) ** 2))

movement_data <- rbind(movement_week1, movement_week2, movement_week3, movement_week4, movement_week5, movement_week6, movement_week7, movement_week8, track_week9)

#write.csv(movement_data, "movement_data.csv")

movement_data
```

# Adding variables to WR play by play data (Errors in information right now) ---------------------------------

## Adding position & distance variables to WR play by play data
```{r}
plays_of_WR <- plays_of_WR |> 
  inner_join(movement_data, by = c("gameId", "playId", "nflId", "displayName"))
```

## Add variable indicating whether or not the player motioned to WR play by play data
```{r}
plays_of_WR <- plays_of_WR |>
  mutate(motioned = ifelse(!is.na(inMotionAtBallSnap), "Yes", "No")) # NAs represent players that didn't motion
```

## Add variable indicating whether or not the WR was elligible to motion to WR play by play data
```{r}
# plays of players where they were set at the line
plays_of_WR <- plays_of_WR |> 
  mutate(set_at_line = ifelse(!is.na(motionSinceLineset), "Yes", "No")) # NAs represent players that were not set at line
```

## Adding how many players motioned on a given play 

```{r}
plays_with_motion <- predicting_plays |> 
  count(gameId, playId, motioned) |> 
  filter(motioned == "Yes") |> 
  select(-motioned) |> 
  rename(num_players_motioned = n) 

plays_of_WR <- plays_of_WR |> 
  left_join(plays_with_motion) |> 
  mutate(num_players_motioned = ifelse(is.na(num_players_motioned), 0, num_players_motioned))
```


## Adding score differential 

```{r}

home_teams <- games |> 
  select(gameId, homeTeamAbbr)

plays_of_WR <- plays_of_WR |> 
  inner_join(home_teams) |> 
  mutate(home_team = ifelse(possessionTeam == homeTeamAbbr, "Yes", "No"),
         score_differential = ifelse(possessionTeam == homeTeamAbbr, preSnapHomeScore - preSnapVisitorScore, preSnapVisitorScore - preSnapHomeScore)) 
```

```{r}
#write.csv(plays_of_WR, "wr_model_data.csv")
```

# Adding variables to all offensive players play by play data --------------

```{r}
predicting_plays <- predicting_plays |> 
  inner_join(movement_data)
```

