---
title: "Appendix"
output: html_document
---

# Appendix ----------------------------------------------------------------------------

# Identifying which player position to focus on in our analysis --------------------------

```{r}
# identify all players (by nflID) that play WR 
WR <- players |> 
  filter(position == "WR") |> 
  distinct(nflId)

# identify all players that ran at least one route 
route_runners <- player_play |> 
  drop_na(routeRan) |> 
  distinct(nflId)

# identify which players were not WR that ran a route 
Non_WR_route_runners <- anti_join(route_runners, WR)
  
# identify which positions ran routes that were not WR and how often - notice there are QB, FB, TE, and RB as well
left_join(Non_WR_route_runners, players) |> 
  count(position)

# find out how many players play each position
players |> 
  count(position)
```

# Exploring the NAs left ------------------------------- 

```{r}
# See how many plays now have NAs for routes ran - 15,393
plays_of_WR |> 
  filter(routeRan == "NO ROUTE")

# There are no plays that were blown dead because of penalty
plays_of_WR |> 
  count(playNullifiedByPenalty)
```

## Figuring out when they might be blocking

```{r}
# All plays where the WR blocked a player- on 11 of them the WR still ran a route
plays_of_WR |> 
  filter(!is.na(blockedPlayerNFLId1))
```

```{r}
# Find out which routes have at least one WR that ran that route and one WR that did not run a route - makes sense that screen is one of the highest 
plays_of_WR |> 
  inner_join(no_route_ran) |> 
  filter(routeRan != "NO ROUTE") |> 
  count(routeRan)
```

## Identifying receivers the did not run routes when screen route was run
```{r}
screen_plays <- plays_of_WR |> 
  filter(routeRan == "SCREEN") |> 
  distinct(gameId, playId)

plays_of_WR |> 
  inner_join(screen_plays)|> 
  filter(routeRan == "NO ROUTE")|> 
  group_by(gameId, playId) |> 
  summarise(num_receivers_no_route = n()) |> 
  ungroup() |> 
  count(num_receivers_no_route)
```

## Figuring out if receivers are running routes on sacks
```{r}
plays_of_WR |> 
  filter(passResult == "S") |> 
  count(routeRan)
```

# Exploring Formations ------------------------------

## Jumbo 
```{r}

# filtering for plays where the formation was Jumbo
jumbo <- plays_of_WR |> 
  filter(offenseFormation == "JUMBO"
         ) |> 
  distinct(gameId, playId)

jumbo

# answered questions like - (1) are plays with off formation jumbo always runs (no) (2) how often in the jumbo formation do receivers run routes (3) how often do receivers block on jumbo plays
plays_of_WR |> 
  inner_join(jumbo) |> 
  filter(!is.na(passLength)) |> 
  #filter(routeRan != "NO ROUTE") |> 
  #filter(is.na(blockedPlayerNFLId1)) |> 
  #count(gameId, playId) |> 
  #count(n)

```

## Wildcat
```{r}

# filtering for plays where the formation was wildcat
wildcat <- plays_of_WR |> 
  filter(offenseFormation == "WILDCAT"
         ) |> 
  distinct(gameId, playId)

wildcat

# how many WR do not run routes when in the wildcat formation - can not remove wildcats because receivers are still running routes
plays_of_WR |> 
  inner_join(wildcat) |> 
  filter(routeRan != "NO ROUTE") |> 
  count(gameId, playId) |> 
  count(n) |> 
  rename(num_receivers = n,
         num_occurrences = nn)

```

# Figuring Out Differences / Relationships Between `motionSinceLineset` & `inMotionAtBallSnap` ------------

```{r}
plays_of_WR |> 
  count(inMotionAtBallSnap)



plays_of_WR |> 
  filter(motionSinceLineset == FALSE | is.na(motionSinceLineset),
        inMotionAtBallSnap == TRUE)

plays_of_WR |> 
  count(shiftSinceLineset)

```