---
title: "Cleaned Data"
author: "Belle Schmidt"
date: "2025-01-02"
output: html_document
---

# Set Up ---------------------

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
library(stringr)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_8.csv")
tracking_week9 <- read.csv("tracking_week_9.csv")
```

# Only including players on offense ------------------------------------------

```{r}
possessionTeam <- plays |> 
  select(gameId, playId, possessionTeam, quarter, down, yardsToGo, offenseFormation)

teamName <- player_play |> 
  select(gameId, playId, nflId, teamAbbr, routeRan, motionSinceLineset, inMotionAtBallSnap, shiftSinceLineset)

players_on_offense <- teamName |> 
  inner_join(possessionTeam) |> 
  filter(teamAbbr == possessionTeam) |> 
  select(-teamAbbr, -possessionTeam)
```

```{r}
tracking_week1 <- inner_join(tracking_week1, players_on_offense) 

tracking_week2 <- inner_join(tracking_week2, players_on_offense)

tracking_week3 <- inner_join(tracking_week3, players_on_offense)

tracking_week4 <- inner_join(tracking_week4, players_on_offense)

tracking_week5 <- inner_join(tracking_week5, players_on_offense)

tracking_week6 <- inner_join(tracking_week6, players_on_offense)

tracking_week7 <- inner_join(tracking_week7, players_on_offense)

tracking_week8 <- inner_join(tracking_week8, players_on_offense)

tracking_week9 <- inner_join(tracking_week9, players_on_offense)
```

# Excluding After Snap Frames ---------------

```{r}
tracking_week1 <- tracking_week1 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week2 <- tracking_week2 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week3 <- tracking_week3 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week4 <- tracking_week4 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week5 <- tracking_week5 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week6 <- tracking_week6 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week7 <- tracking_week7 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week8 <- tracking_week8 |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week9 <- tracking_week9 |> 
  filter(frameType != "AFTER_SNAP") 
```


# Excluding Certain Plays -------------------

## Removing QB Spikes
```{r}
# Plays where the ball is not spiked 
non_spikes <- plays |> 
  filter(qbSpike == FALSE | is.na(qbSpike)) |> 
  select(gameId, playId)

tracking_week1 <- inner_join(tracking_week1, non_spikes)

tracking_week2 <- inner_join(tracking_week2, non_spikes)

tracking_week3 <- inner_join(tracking_week3, non_spikes)

tracking_week4 <- inner_join(tracking_week4, non_spikes)

tracking_week5 <- inner_join(tracking_week5, non_spikes)

tracking_week6 <- inner_join(tracking_week6, non_spikes)

tracking_week7 <- inner_join(tracking_week7, non_spikes)

tracking_week8 <- inner_join(tracking_week8, non_spikes)

tracking_week9 <- inner_join(tracking_week9, non_spikes)
```

## Removing QB Kneel Plays

```{r}
kneels <- plays |> 
  filter(qbKneel == 1) |> 
  select(gameId, playId)

tracking_week1 <- anti_join(tracking_week1, kneels)

tracking_week2 <- anti_join(tracking_week2, kneels)

tracking_week3 <- anti_join(tracking_week3, kneels)

tracking_week4 <- anti_join(tracking_week4, kneels)

tracking_week5 <- anti_join(tracking_week5, kneels)

tracking_week6 <- anti_join(tracking_week6, kneels)

tracking_week7 <- anti_join(tracking_week7, kneels)

tracking_week8 <- anti_join(tracking_week8, kneels)

tracking_week9 <- anti_join(tracking_week9, kneels)
```

## Removing QB Sneaks
```{r}
# remove QB sneaks from the data. This is okay to remove because QB sneaks are very predictable plays so the defense already knows the receiver isn't running a route. - There are still 40,438 receiver plays
non_sneaks <- plays |> 
  filter(is.na(qbSneak) | qbSneak == FALSE) |> 
  select(gameId, playId)

tracking_week1 <- inner_join(tracking_week1, non_sneaks)

tracking_week2 <- inner_join(tracking_week2, non_sneaks)

tracking_week3 <- inner_join(tracking_week3, non_sneaks)

tracking_week4 <- inner_join(tracking_week4, non_sneaks)

tracking_week5 <- inner_join(tracking_week5, non_sneaks)

tracking_week6 <- inner_join(tracking_week6, non_sneaks)

tracking_week7 <- inner_join(tracking_week7, non_sneaks)

tracking_week8 <- inner_join(tracking_week8, non_sneaks)

tracking_week9 <- inner_join(tracking_week9, non_sneaks)
```

# Dealing with play where two frames have "line_set" moment  ------------------------

```{r}
# Only occurred on one play in week 1
plays_with_new_set_line_week1 <- tracking_week1 |> 
  group_by(gameId, playId, nflId) |> 
  filter(event == "line_set") |> 
  summarise(num_times_set = n()) |> 
  ungroup() |> 
  filter(num_times_set == 2) |> 
  distinct(gameId, playId)

set_line_change_week1 <- tracking_week1 |> 
  inner_join(plays_with_new_set_line_week1) |> 
  distinct(gameId, playId)

set_line_change_week1

tracking_week1 <- tracking_week1 |>
  mutate(event = ifelse(gameId == 2022091103 & playId == 4434 & frameId == 4, "second_line_set", event))
```

# Identifying which players are included in the receiver alignment  --------------------

## Setting up the data to determine which players are included in the receiver alignment 

### Adding official position as a variable
```{r}
official_position <- players |> 
  select(nflId, position) |> 
  rename(official_position = position)

tracking_week1 <- inner_join(tracking_week1, official_position)

tracking_week2 <- inner_join(tracking_week2, official_position)

tracking_week3 <- inner_join(tracking_week3, official_position)

tracking_week4 <- inner_join(tracking_week4, official_position)

tracking_week5 <- inner_join(tracking_week5, official_position)

tracking_week6 <- inner_join(tracking_week6, official_position)

tracking_week7 <- inner_join(tracking_week7, official_position)

tracking_week8 <- inner_join(tracking_week8, official_position)

tracking_week9 <- inner_join(tracking_week9, official_position)
```

### Adding variables indicating how many receivers lined up on the each side of the line, how many total receivers, and how far the line of scrimmage is from the end zone they are trying to score in 
```{r}
# add the absolute yard line (ranging 0 to 100),  number of receivers, and number of receivers on the left and right side variables to the tracking data (created receiver variables)
important_variables <- plays |> 
  mutate(num_receivers_one_side = as.numeric(stringr::str_extract(receiverAlignment, "\\d{1}")),
         num_receivers_other_side = as.numeric(stringr::str_extract(receiverAlignment, "(?<=x)[0-9]+")),
         total_receivers_in_alignment = num_receivers_one_side + num_receivers_other_side,
         yardsTillTouchdown = ifelse(possessionTeam != yardlineSide, yardlineNumber, 100 - yardlineNumber),  # makes it so the yard line variable is how far they are from their the end zone they are scoring in 
         yardsTillTouchdown = ifelse(yardlineNumber == 50, 50, yardsTillTouchdown),
         absoluteYardlineNumber = absoluteYardlineNumber - 10) |> 
  select(gameId, playId, yardsTillTouchdown, total_receivers_in_alignment, num_receivers_one_side, num_receivers_other_side, preSnapHomeScore, preSnapVisitorScore)

tracking_week1 <- inner_join(tracking_week1, important_variables)

tracking_week2 <- inner_join(tracking_week2, important_variables)

tracking_week3 <- inner_join(tracking_week3, important_variables)

tracking_week4 <- inner_join(tracking_week4, important_variables)

tracking_week5 <- inner_join(tracking_week5, important_variables)

tracking_week6 <- inner_join(tracking_week6, important_variables)

tracking_week7 <- inner_join(tracking_week7, important_variables)

tracking_week8 <- inner_join(tracking_week8, important_variables)

tracking_week9 <- inner_join(tracking_week9, important_variables)

```


### Creating a variable that indicates how far the players are from the end zone they are trying to score in (rounded and not) 
```{r}

adding_player_x_position <- function(week_num){
  week_num |> 
  mutate(yardlines_to_go = ifelse(playDirection == "left", round(x - 10), NA),
         exact_yards_to_go = ifelse(playDirection == "left", x - 10, NA),
         yardlines_to_go = ifelse(playDirection == "right", 100 - round(x - 10), yardlines_to_go), 
         exact_yards_to_go = ifelse(playDirection == "right", 100 - (x - 10), exact_yards_to_go),
         yardlines_to_go = ifelse(frameType == "SNAP" & (yardsTillTouchdown > yardlines_to_go), yardsTillTouchdown, yardlines_to_go))  # all of these were off by less than a yard
}

tracking_week1 <- adding_player_x_position(tracking_week1)

tracking_week2 <- adding_player_x_position(tracking_week2)

tracking_week3 <- adding_player_x_position(tracking_week3)

tracking_week4 <- adding_player_x_position(tracking_week4)

tracking_week5 <- adding_player_x_position(tracking_week5)

tracking_week6 <- adding_player_x_position(tracking_week6)

tracking_week7 <- adding_player_x_position(tracking_week7)

tracking_week8 <- adding_player_x_position(tracking_week8)

tracking_week9 <- adding_player_x_position(tracking_week9)
```

### Indicating which player number they are (1 means furthest left and 11 means furthest right - in terms of direction they are facing)
```{r}
player_numbers <- function(week_num){
  week_num |> 
    group_by(gameId, playId, frameId) |> 
    arrange(gameId, playId, frameId, y) |>
    mutate(player_number = ifelse(playDirection == "right", 12 - row_number(), row_number())) |> 
    ungroup() |> 
    select(gameId, playId, nflId, frameId, player_number)
  }
 
tracking_week1 <- inner_join(tracking_week1, player_numbers(tracking_week1))

tracking_week2 <- inner_join(tracking_week2, player_numbers(tracking_week2))

tracking_week3 <- inner_join(tracking_week3, player_numbers(tracking_week3))

tracking_week4 <- inner_join(tracking_week4, player_numbers(tracking_week4))

tracking_week5 <- inner_join(tracking_week5, player_numbers(tracking_week5))

tracking_week6 <- inner_join(tracking_week6, player_numbers(tracking_week6))

tracking_week7 <- inner_join(tracking_week7, player_numbers(tracking_week7))

tracking_week8 <- inner_join(tracking_week8, player_numbers(tracking_week8))

tracking_week9 <- inner_join(tracking_week9, player_numbers(tracking_week9))
```

# IMPORTANT NOTE: This is assuming that the running back in the backfield is NOT included in the receiver alignment

### Initially trying to determine if player is in left or right alignment
```{r}
receiver_alignment <- function(num_week) {
  num_week |> 
    filter(frameType == "SNAP") |> 
    mutate(receiver_in_left_alignment = ifelse(player_number %in% c(1, 2, 3) & (!is.na(routeRan) | official_position %in% c("TE", "WR")), TRUE, FALSE),
         receiver_in_right_alignment = ifelse(player_number %in% c(9, 10, 11) & (!is.na(routeRan) | official_position %in% c("TE", "WR")), TRUE, FALSE)) |> 
    select(gameId, playId, nflId, receiver_in_left_alignment, receiver_in_right_alignment)
}

receiver_alignment_week1 <- receiver_alignment(tracking_week1)

receiver_alignment_week2 <- receiver_alignment(tracking_week2)

receiver_alignment_week3 <- receiver_alignment(tracking_week3)

receiver_alignment_week4 <- receiver_alignment(tracking_week4)

receiver_alignment_week5 <- receiver_alignment(tracking_week5)

receiver_alignment_week6 <- receiver_alignment(tracking_week6)

receiver_alignment_week7 <- receiver_alignment(tracking_week7)

receiver_alignment_week8 <- receiver_alignment(tracking_week8)

receiver_alignment_week9 <- receiver_alignment(tracking_week9)


tracking_week1 <- inner_join(tracking_week1, receiver_alignment_week1)

tracking_week2 <- inner_join(tracking_week2, receiver_alignment_week2)

tracking_week3 <- inner_join(tracking_week3, receiver_alignment_week3)

tracking_week4 <- inner_join(tracking_week4, receiver_alignment_week4)

tracking_week5 <- inner_join(tracking_week5, receiver_alignment_week5)

tracking_week6 <- inner_join(tracking_week6, receiver_alignment_week6)

tracking_week7 <- inner_join(tracking_week7, receiver_alignment_week7)

tracking_week8 <- inner_join(tracking_week8, receiver_alignment_week8)

tracking_week9 <- inner_join(tracking_week9, receiver_alignment_week9)
```

### Adding variable to count how many receivers are currently in left & right alignment 

```{r}
num_receivers_left <- function(receivers_left) {
  receivers_left |> 
    filter(frameType == "SNAP",
           receiver_in_left_alignment == TRUE) |> 
    count(gameId, playId) |> 
    rename(num_receivers_left = n)
}

number_receivers_left_week1 <- num_receivers_left(tracking_week1)

number_receivers_left_week2 <- num_receivers_left(tracking_week2)

number_receivers_left_week3 <- num_receivers_left(tracking_week3)

number_receivers_left_week4 <- num_receivers_left(tracking_week4)

number_receivers_left_week5 <- num_receivers_left(tracking_week5)

number_receivers_left_week6 <- num_receivers_left(tracking_week6)

number_receivers_left_week7 <- num_receivers_left(tracking_week7)

number_receivers_left_week8 <- num_receivers_left(tracking_week8)

number_receivers_left_week9 <- num_receivers_left(tracking_week9)

tracking_week1 <- inner_join(tracking_week1, number_receivers_left_week1)

tracking_week2 <- inner_join(tracking_week2, number_receivers_left_week2)

tracking_week3 <- inner_join(tracking_week3, number_receivers_left_week3)

tracking_week4 <- inner_join(tracking_week4, number_receivers_left_week4)

tracking_week5 <- inner_join(tracking_week5, number_receivers_left_week5)

tracking_week6 <- inner_join(tracking_week6, number_receivers_left_week6)

tracking_week7 <- inner_join(tracking_week7, number_receivers_left_week7)

tracking_week8 <- inner_join(tracking_week8, number_receivers_left_week8)

tracking_week9 <- inner_join(tracking_week9, number_receivers_left_week9)
```

```{r}
num_receivers_right <- function(receivers_right) {
  receivers_right |> 
    filter(frameType == "SNAP",
           receiver_in_right_alignment == TRUE) |> 
    count(gameId, playId) |> 
    rename(num_receivers_right = n)
}

number_receivers_right_week1 <- num_receivers_right(tracking_week1)

number_receivers_right_week2 <- num_receivers_right(tracking_week2)

number_receivers_right_week3 <- num_receivers_right(tracking_week3)

number_receivers_right_week4 <- num_receivers_right(tracking_week4)

number_receivers_right_week5 <- num_receivers_right(tracking_week5)

number_receivers_right_week6 <- num_receivers_right(tracking_week6)

number_receivers_right_week7 <- num_receivers_right(tracking_week7)

number_receivers_right_week8 <- num_receivers_right(tracking_week8)

number_receivers_right_week9 <- num_receivers_right(tracking_week9)

tracking_week1 <- inner_join(tracking_week1, number_receivers_right_week1)

tracking_week2 <- inner_join(tracking_week2, number_receivers_right_week2)

tracking_week3 <- inner_join(tracking_week3, number_receivers_right_week3)

tracking_week4 <- inner_join(tracking_week4, number_receivers_right_week4)

tracking_week5 <- inner_join(tracking_week5, number_receivers_right_week5)

tracking_week6 <- inner_join(tracking_week6, number_receivers_right_week6)

tracking_week7 <- inner_join(tracking_week7, number_receivers_right_week7)

tracking_week8 <- inner_join(tracking_week8, number_receivers_right_week8)

tracking_week9 <- inner_join(tracking_week9, number_receivers_right_week9)

# 14,561 total plays
```

# Identifying which plays have the correct receivers in alignment --------------------

```{r}
finished_plays_playdID <- function(receiver_alignment_num_week) {
  receiver_alignment_num_week |> 
    filter(num_receivers_left + num_receivers_right == total_receivers_in_alignment)
}


finished_plays_week1 <- finished_plays_playdID(tracking_week1)

finished_plays_week2 <- finished_plays_playdID(tracking_week2)

finished_plays_week3 <- finished_plays_playdID(tracking_week3)

finished_plays_week4 <- finished_plays_playdID(tracking_week4)

finished_plays_week5 <- finished_plays_playdID(tracking_week5)

finished_plays_week6 <- finished_plays_playdID(tracking_week6)

finished_plays_week7 <- finished_plays_playdID(tracking_week7)

finished_plays_week8 <- finished_plays_playdID(tracking_week8)

finished_plays_week9 <- finished_plays_playdID(tracking_week9)
```


```{r}
finished_plays <- rbind(finished_plays_week1, finished_plays_week2, finished_plays_week3, finished_plays_week4, finished_plays_week5, finished_plays_week6, finished_plays_week7, finished_plays_week8, finished_plays_week9)

# there are 11,098 distinct plays
finished_plays |> 
  distinct(gameId, playId)

messed_up <- finished_plays |> 
  filter(num_receivers_left != num_receivers_other_side & num_receivers_left != num_receivers_one_side) 

# 184 messed up plays
messed_up |> 
  distinct(gameId, playId)

finished_plays <- anti_join(finished_plays, messed_up) 
  
# 10,914 plays completely finished 
finished_plays |> 
  distinct(gameId, playId)
```

## Fixing Messed Up Plays (COME BACK TO THIS)

```{r}
messed_up |> 
  filter(frameType == "SNAP",
         receiver_in_left_alignment == TRUE | receiver_in_right_alignment == TRUE) |> 
  arrange(gameId, playId, player_number) 
  
```


# Identifying plays with too many receivers in alignment -------------------------

```{r}
too_many_receivers_playdID <- function(receiver_alignment_num_week) {
  receiver_alignment_num_week |> 
    filter(num_receivers_left + num_receivers_right > total_receivers_in_alignment)
}


too_many_receivers_week1 <- too_many_receivers_playdID(tracking_week1)

too_many_receivers_week2 <- too_many_receivers_playdID(tracking_week2) 

too_many_receivers_week3 <- too_many_receivers_playdID(tracking_week3)

too_many_receivers_week4 <- too_many_receivers_playdID(tracking_week4)

too_many_receivers_week5 <- too_many_receivers_playdID(tracking_week5)

too_many_receivers_week6 <- too_many_receivers_playdID(tracking_week6)

too_many_receivers_week7 <- too_many_receivers_playdID(tracking_week7)

too_many_receivers_week8 <- too_many_receivers_playdID(tracking_week8)

too_many_receivers_week9 <- too_many_receivers_playdID(tracking_week9)
```

```{r}
too_many_receiver_plays <- rbind(too_many_receivers_week1, too_many_receivers_week2, too_many_receivers_week3, too_many_receivers_week4, too_many_receivers_week5, too_many_receivers_week6, too_many_receivers_week7, too_many_receivers_week8, too_many_receivers_week9)

# there are 2,227 distinct plays
too_many_receiver_plays |> 
  filter(frameType == "SNAP") |> 
  distinct(gameId, playId)
```

## Identifying which side is messed up 

```{r}
messed_up_left <- too_many_receiver_plays |> 
  filter(num_receivers_left != num_receivers_other_side & num_receivers_left != num_receivers_one_side) 

# guaranteed messed up left - 1,029
messed_up_left_plays <- messed_up_left |> 
  distinct(gameId, playId)

messed_up_right <- too_many_receiver_plays |> 
  filter(num_receivers_right != num_receivers_other_side & num_receivers_right != num_receivers_one_side) 

# guaranteed messed up right - 894
messed_up_right_plays <- messed_up_right |> 
  distinct(gameId, playId)


messed_up_side_unsure <- too_many_receiver_plays |> 
  anti_join(messed_up_left_plays)

messed_up_side_unsure <- messed_up_side_unsure |> 
  anti_join(messed_up_right_plays)

# 1,198 plays where messed up side is unsure
messed_up_side_unsure |> 
  distinct(gameId, playId)

# Note: It makes sense that the number of plays don't add up to the total number of plays messed up because some plays might have left and right side messed up 

```

## Identifying which players are not actually receivers in alignment ----------------

### Plays with only one receiver in alignment ------------------------------

```{r}
# identifying plays where only one receiver was in alignment 
one_receiver_in_alignment <- too_many_receiver_plays |> 
  filter(total_receivers_in_alignment == 1)

# creates a function that fixes the receiver alignment 
one_receiver_in_alignment_receivers_side <- function(number, fixing_side, other_side){
  
  # first starts by finding a way to identify the receivers in alignment
  start <- too_many_receiver_plays |>
    inner_join(one_receiver_in_alignment) |> 
    filter(frameType == "SNAP",
           player_number == number,
           official_position %in% c("WR", "TE"))
  
  # plays where the receivers have been identified using the method in start
  plays <- start |> 
    distinct(gameId, playId)
  
  # identified receivers using the method in start
  receivers <- start |> 
  distinct(gameId, playId, displayName) |> 
  mutate(!!paste0("receiver_in_", fixing_side, "_alignment_new") := TRUE)

# fixed the plays where the one receiver is in alignment 
  final_data <- too_many_receiver_plays |> 
  inner_join(plays) |> 
  left_join(receivers) |> 
  mutate(!!paste0("receiver_in_", fixing_side, "_alignment") := ifelse(is.na(!!sym(paste0("receiver_in_", fixing_side, "_alignment_new"))), FALSE, TRUE),
         !!paste0("num_receivers_", fixing_side) := 1,
         !!paste0("num_receivers_", other_side) := 0) |> 
  select(- !!sym(paste0("receiver_in_", fixing_side, "_alignment_new")))
  
  return(final_data)
}
```

#### Fixed one total receiver plays
```{r}
finished_total_one_receiver_left <- one_receiver_in_alignment_receivers_side(1, "left", "right")

finished_total_one_receiver_right <- one_receiver_in_alignment_receivers_side(1, "right", "left")

finished_one_receiver_plays <- rbind(finished_total_one_receiver_left, finished_total_one_receiver_right)
```


### Plays with two receivers in alignment; one on each side ----------------------------

```{r}
fixing_receivers <- function(data, t, l, r) {
  data |> 
  filter(total_receivers_in_alignment == t, num_receivers_one_side == l,
         num_receivers_other_side == r,
         frameType == "SNAP")
}

total_2_left_1_right_1 <- fixing_receivers(too_many_receiver_plays, 2, 1, 1)

# there are 48 plays with 1 receiver on each side
total_2_left_1_right_1_plays <- total_2_left_1_right_1 |> 
    distinct(gameId, playId)

# 27 of them do not have one receiver identified on the left side
Total_2_Left_1_left_messed_up <- total_2_left_1_right_1 |> 
  filter(num_receivers_left != 1) |> 
  distinct(gameId, playId)

# 23 of the 27 have WR in the 1 position
total_2_left_1_right_1_position1 <- total_2_left_1_right_1 |> 
  inner_join(Total_2_Left_1_left_messed_up) |> 
  filter(player_number == 1,
         official_position %in% c("WR")) |> 
    distinct(gameId, playId) 

# 40 of them do not have one receiver identified on the right side
Total_2_Left_1_right_messed_up <- total_2_left_1_right_1 |> 
  filter(num_receivers_right != 1) |> 
  distinct(gameId, playId)
```

#### Fixing the 23 plays with WR in position 1

```{r}
# these are the left receivers in those plays
fix_one_receiver_left_part1 <- total_2_left_1_right_1 |> 
  filter(player_number == 1,
         official_position == "WR") |> 
    distinct(gameId, playId, displayName) |> 
    mutate(receiver_in_left_alignment_new = TRUE)

# Fixes the left side of the plays with WR in position 1 at snap
total_two_left_one_plays_fixed_part1 <- too_many_receiver_plays |> 
  inner_join(total_2_left_1_right_1_position1) |> 
  left_join(fix_one_receiver_left_part1) |> 
  mutate(receiver_in_left_alignment = ifelse(is.na(receiver_in_left_alignment_new), FALSE, TRUE),
         num_receivers_left = 1) |> 
  select(- receiver_in_left_alignment_new) 

# Seven of the plays are completely fixed; the others need the right side fixed
finished_total_two_left_one_right_one_part1 <- total_two_left_one_plays_fixed_part1 |> 
  anti_join(Total_2_Left_1_right_messed_up)

# Fixing right side of the plays where the left receiver is a WR in position 1
fix_one_receiver_right_part1 <- total_two_left_one_plays_fixed_part1 |> 
  anti_join(finished_total_two_left_one_right_one_part1) |> 
  filter(frameType == "SNAP",
         player_number == 11,
         official_position == "WR") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_right_alignment_new = TRUE)

# Now all of the 23 plays with a WR in position 1 are finished
finished_total_two_left_one_right_one_part2 <- total_two_left_one_plays_fixed_part1 |> 
  anti_join(finished_total_two_left_one_right_one_part1) |> 
  left_join(fix_one_receiver_right_part1) |> 
  mutate(receiver_in_right_alignment = ifelse(is.na(receiver_in_right_alignment_new), FALSE, TRUE),
         num_receivers_right = 1) |> 
  select(- receiver_in_right_alignment_new)

```

#### Fixing the 4 plays that did not have WRs in the 1 position BUT had left side messed up

```{r}
# one way of identifying the receiver in alignment 
fix_play_w_WR <- total_2_left_1_right_1 |> 
  inner_join(Total_2_Left_1_left_messed_up) |> 
  anti_join(total_2_left_1_right_1_position1)|> 
  filter(player_number < 7 & official_position == "WR") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_left_alignment_new = TRUE)

# second way to identify the receiver in alignment
fix_plays_w_TE <- total_2_left_1_right_1 |> 
  inner_join(Total_2_Left_1_left_messed_up) |> 
  anti_join(total_2_left_1_right_1_position1) |> 
  anti_join(fix_play_w_WR, join_by(gameId, playId)) |> 
  filter(player_number == 1 & official_position == "TE") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_left_alignment_new = TRUE)

# combining left receivers
fix_left_side_of_remaining_two_total_one_each_side <- rbind(fix_play_w_WR, fix_plays_w_TE)

# fixing which player is in left alignment
fixed_left_side <- too_many_receiver_plays |>
  inner_join(Total_2_Left_1_left_messed_up) |> 
  anti_join(total_2_left_1_right_1_position1) |> 
  left_join(fix_left_side_of_remaining_two_total_one_each_side) |> 
  mutate(receiver_in_left_alignment = ifelse(is.na(receiver_in_left_alignment_new), FALSE, TRUE),
         num_receivers_left = 1) |> 
  select(- receiver_in_left_alignment_new) 

# plays that already had the correct receiver in alignment on the right side
finished_total_two_left_one_right_one_part3 <- fixed_left_side |> 
  anti_join(Total_2_Left_1_right_messed_up)

# Fixing the right side for plays that did not already have the right side correct -------- 
# one way of identifying the receiver in right alignment 
fix_plays_w_receivers_right <- fixed_left_side |> 
  anti_join(finished_total_two_left_one_right_one_part3) |> 
  filter(frameType == "SNAP",
         player_number > 6 & official_position == "WR") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_right_alignment_new = TRUE)

# second way of identifying the right receiver in alignment
fix_play_w_TE_right <- fixed_left_side |> 
  anti_join(finished_total_two_left_one_right_one_part3) |> 
  anti_join(fix_plays_w_receivers_right, join_by(gameId, playId)) |> 
  filter(frameType == "SNAP",
         player_number == 11 & official_position == "TE") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_right_alignment_new = TRUE)
  
# combining right receivers in alignment 
fix_right_side_of_remaining_two_total_one_each_side <- rbind(fix_plays_w_receivers_right, fix_play_w_TE_right)

# Right side fixed so all plays are now finished
finished_total_two_left_one_right_one_part4 <- fixed_left_side |> 
  anti_join(finished_total_two_left_one_right_one_part3) |> 
  left_join(fix_right_side_of_remaining_two_total_one_each_side) |> 
  mutate(receiver_in_right_alignment = ifelse(is.na(receiver_in_right_alignment_new), FALSE, TRUE),
         num_receivers_right = 1) |> 
  select(- receiver_in_right_alignment_new)

```

#### Fixing plays that only had the right side messed up

```{r}
# 11 of 21 of the remaining plays have WR in the 11 position
total_2_left_1_right_1 |> 
  anti_join(Total_2_Left_1_left_messed_up) |> 
  filter(frameType == "SNAP",
         player_number == 11,
         official_position == "WR") |> 
  distinct(gameId, playId, displayName) |> 
  mutate(receiver_in_right_alignment_new = TRUE)

# Now all of the 23 plays with a WR in position 1 are finished
finished_total_two_left_one_right_one_part2 <- total_two_left_one_plays_fixed_part1 |> 
  anti_join(finished_total_two_left_one_right_one_part1) |> 
  left_join(fix_one_receiver_right_part1) |> 
  mutate(receiver_in_right_alignment = ifelse(is.na(receiver_in_right_alignment_new), FALSE, TRUE),
         num_receivers_right = 1) |> 
  select(- receiver_in_right_alignment_new)
```


#### All the plays fixed that had two total receivers, one each side, and too many identified receivers originally
```{r}
finished_one_left_one_right_receiver_plays <- rbind(finished_total_two_left_one_right_one_part1, finished_total_two_left_one_right_one_part2, finished_total_two_left_one_right_one_part3, finished_total_two_left_one_right_one_part4)
```

### Plays with two receivers in alignment; two on left side and zero on right ------------

```{r}
total_2_left_2_right_0 <- fixing_receivers(too_many_receiver_plays, 2, 2, 0)

total_2_left_2_right_0
```


# Identifying plays with not enough receivers in alignment -------------------------

```{r}
not_enough_receivers_playdID <- function(receiver_alignment_num_week) {
  receiver_alignment_num_week |> 
    filter(num_receivers_left + num_receivers_right < total_receivers_in_alignment)
}


not_enough_receivers_week1 <- not_enough_receivers_playdID(tracking_week1)

not_enough_receivers_week2 <- not_enough_receivers_playdID(tracking_week2)

not_enough_receivers_week3 <- not_enough_receivers_playdID(tracking_week3)

not_enough_receivers_week4 <- not_enough_receivers_playdID(tracking_week4)

not_enough_receivers_week5 <- not_enough_receivers_playdID(tracking_week5)

not_enough_receivers_week6 <- not_enough_receivers_playdID(tracking_week6)

not_enough_receivers_week7 <- not_enough_receivers_playdID(tracking_week7)

not_enough_receivers_week8 <- not_enough_receivers_playdID(tracking_week8)

not_enough_receivers_week9 <- not_enough_receivers_playdID(tracking_week9)
```

```{r}
not_enough_receiver_plays <- rbind(not_enough_receivers_week1, not_enough_receivers_week2, not_enough_receivers_week3, not_enough_receivers_week4, not_enough_receivers_week5, not_enough_receivers_week6, not_enough_receivers_week7, not_enough_receivers_week8, not_enough_receivers_week9)

# there are 1,236 distinct plays
not_enough_receiver_plays |> 
  distinct(gameId, playId)
```

# Adding Variables ------------------------

## Adding Score Differential 

```{r}
home_teams <- games |> 
  select(gameId, homeTeamAbbr)

finished_plays <- finished_plays |> 
  inner_join(home_teams) |> 
  mutate(score_differential = ifelse(club == homeTeamAbbr, preSnapHomeScore - preSnapVisitorScore, preSnapVisitorScore - preSnapHomeScore)) |> 
  select(-homeTeamAbbr, -preSnapHomeScore, -preSnapVisitorScore)
```

## Addding a variable `moved` indicating if they moved after their line was set

```{r}
finished_plays <- finished_plays |> 
  mutate(moved = ifelse(shiftSinceLineset == TRUE | inMotionAtBallSnap == TRUE | motionSinceLineset == TRUE, TRUE, FALSE),
         moved = ifelse(is.na(moved), FALSE, TRUE)) 
```

## Adding a variable `num_players_moved` indicating how many players moved on the play

```{r}
num_players_moved <- finished_plays |> 
  filter(moved == TRUE,
         frameType == "SNAP") |> 
  count(gameId, playId) |> 
  rename(num_players_moved = n)

no_movement <- finished_plays |> 
  anti_join(num_players_moved) |> 
  distinct(gameId, playId) |> 
  mutate(num_players_moved = 0)

num_players_moved <- rbind(num_players_moved, no_movement)

finished_plays <- finished_plays |> 
  inner_join(num_players_moved)
```

## Adding variables


### Adding distance from line of scrimmage variable 
```{r}
finished_plays <- finished_plays |> 
  mutate(receiver_in_alignment = ifelse(receiver_in_left_alignment == TRUE | receiver_in_right_alignment == TRUE, TRUE, FALSE)) |> 
  mutate(dist_from_line_of_scrimmage = abs(yardsTillTouchdown - yardlines_to_go))
```

### Adding receiver alignment number
```{r}
finished_plays_more_info <- finished_plays |>  
  filter(receiver_in_alignment == TRUE,
         frameType == "SNAP") |> 
  group_by(gameId, playId) |> 
  arrange(gameId, playId, y) |> 
  mutate(left_or_right = ifelse(receiver_in_left_alignment == TRUE, "left", "right"),
    receiver_number = ifelse(playDirection == "left" & left_or_right == "left", row_number(), NA),
    receiver_number = ifelse(playDirection == "left" & left_or_right == "right", total_receivers_in_alignment - (row_number() - 1), receiver_number),
    receiver_number = ifelse(playDirection == "right" & left_or_right == "left", total_receivers_in_alignment - (row_number() - 1), receiver_number),
    receiver_number = ifelse(playDirection == "right" & left_or_right == "right", row_number(), receiver_number),
    receiver_number = paste(left_or_right, receiver_number)) |> 
  ungroup() |> 
  select(gameId, playId, nflId, receiver_number)

finished_plays_more_info_whole <- inner_join(finished_plays, finished_plays_more_info)

finished_plays_more_info2 <- finished_plays |>
  filter(receiver_in_alignment == FALSE) |> 
  mutate(receiver_number = NA)

finished_plays <- rbind(finished_plays_more_info_whole, finished_plays_more_info2)
  
```

## Replacing NAs for routeRan

### If run or pass option then if the receiver did not run a route routeRan = NO ROUTE

```{r}
run_pass_option <- plays |> 
  select(gameId, playId, pff_runPassOption)

finished_plays <- inner_join(finished_plays, run_pass_option)


finished_plays <- finished_plays |> 
  mutate(routeRan = ifelse(receiver_in_alignment == TRUE & pff_runPassOption == 1 & is.na(routeRan), "NO ROUTE", routeRan)) 

```

### If the play was a hand off or run play and not pff_runPassOption then routeRan = RUN BLOCK

```{r}
finished_plays <- finished_plays |> 
  mutate(routeRan = ifelse((event == "handoff" | event == "run") & pff_runPassOption == 0 & is.na(routeRan), "RUN BLOCK", routeRan))
```

### If it is a pass play and did not run route then NO ROUTE
```{r}
finished_plays |> 
  filter(receiver_in_alignment == TRUE, 
         frameType == "SNAP",
         is.na(routeRan)) 

pass_plays <- plays |> 
  mutate(pass = ifelse(is.na(passResult), FALSE, TRUE)) |> 
  select(gameId, playId, pass)


finished_plays <- inner_join(finished_plays, pass_plays)

finished_plays <- finished_plays |> 
  mutate(routeRan = ifelse(pass == TRUE & is.na(routeRan), "NO ROUTE", routeRan))
```


## Adding Variables for number of receivers left for each play
```{r}
left_players <- finished_plays |>
  filter(frameType == "SNAP",
         receiver_in_left_alignment == TRUE) |> 
  count(gameId, playId) |> 
  rename(num_receivers_left = n)


no_left_players <- finished_plays |> 
  anti_join(left_players) |> 
  distinct(gameId, playId) |> 
  mutate(num_receivers_left = 0)

left_players <- rbind(left_players, no_left_players)

finished_plays <- inner_join(finished_plays, left_players)
```

## Adding Variables for number of receivers right for each play
```{r}
right_players <- finished_plays |>
  filter(frameType == "SNAP",
         receiver_in_right_alignment == TRUE) |> 
  count(gameId, playId) |> 
  rename(num_receivers_right = n)


no_right_players <- finished_plays |> 
  anti_join(right_players) |> 
  distinct(gameId, playId) |> 
  mutate(num_receivers_right = 0)

right_players <- rbind(right_players, no_right_players)

finished_plays <- inner_join(finished_plays, right_players)
```



# Selecting only plays of players we are predicting

```{r}
players_predicting <- finished_plays |> 
  filter(receiver_in_alignment == TRUE,
         frameType %in% c("BEFORE_SNAP", "SNAP"))

players_predicting


write.csv(players_predicting, "players_predicting.csv")
```

