---
title: "Data Cleaning"
output: html_document
---

# Set Up ---------------------

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_9.csv")
```

# Creating WR Play By Play Data For Prediction Plays --------------------------------------------

## Determining What Plays To Include in Route Running Predicition ------------------

### Setting up the data for more exploration
```{r}
# add play information to individual player play by play data
player_play_w_info <- player_play |> 
  inner_join(plays, join_by(gameId, playId))
```

### Decided to focus on only WR 
```{r}
# Players WR Positions - Include only position of each player, their name, and their NFLId
positions <- players |> 
  filter(position == "WR") |> 
  select(nflId, displayName, position)

# All play information of WRs that ran at least one route - there are 40,940
plays_of_WR <- inner_join(player_play_w_info, positions)
```

### Finding the distribution of the routes ran - it is a little worrisome that so many routes are NAs. There is also a very large disproportion each route is run
```{r}
# The number of times and the proportion of the time each route is run by WR or TEs 
plays_of_WR |> 
  group_by(routeRan) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```

## Figuring out what some of the NAs could mean -------------------

```{r}
# Plays where at least one receiver or tight end did not run a route - there are 6,874 plays
no_route_ran <- plays_of_WR |> 
  filter(is.na(routeRan)) |> 
  distinct(gameId, playId)
```

### Removing Deadplays
```{r}
# Plays where the ball is spiked or the QB kneels - there are 188
deadplays <- plays |> 
  filter(is.na(offenseFormation))

# removes kneeling and spike plays - now there are 6,690 plays
no_route_ran <- anti_join(no_route_ran, deadplays)


# removes kneeling and spike plays from WR play by play info - there are now 40,690 observations

plays_of_WR <- anti_join(plays_of_WR, deadplays, join_by(gameId, playId))
```

### Removing QB Sneaks
```{r}
# remove QB sneaks from the data. This is okay to remove because QB sneaks are very predictable plays so the defense already knows the receiver isn't running a route. - There are still 40,438 receiver plays
plays_of_WR <- plays_of_WR |> 
  filter(is.na(qbSneak) | qbSneak == FALSE)

# Creating a list of plays where qb sneaks are not included
valid_plays <- plays_of_WR |> 
  distinct(gameId, playId)

# Remove QB sneaks from the list of plays where at least one WR with an NA route - 6,571 plays
no_route_ran <- no_route_ran|> 
  inner_join(valid_plays)

```

### Replacing NAs with saying no route ran 
```{r}
plays_of_WR <- plays_of_WR |> 
  mutate(routeRan = replace_na(routeRan, "NO ROUTE")) |> 
  select(gameId:teamAbbr, displayName, position, inMotionAtBallSnap:routeRan, playDescription:preSnapVisitorScore, absoluteYardlineNumber, offenseFormation:playClockAtSnap)
```

```{r}
#write.csv(plays_of_WR, "wr_plays.csv")
```



# Creating play by play data of all offensive players on prediction plays -------------------

### Player data of all players 
```{r}
# All the distinct plays involving WR
plays_w_WR <- plays_of_WR |> 
  distinct(gameId, playId)

# All the plays we are going to do prediction for
predicting_plays <- player_play_w_info |> 
  inner_join(plays_w_WR) |> 
  filter(possessionTeam == teamAbbr)


# including only offensive variables
predicting_plays <- predicting_plays |> 
  select(gameId:teamAbbr, inMotionAtBallSnap:routeRan, playDescription:preSnapVisitorScore, absoluteYardlineNumber, offenseFormation:playClockAtSnap)

```

```{r}
# Players Positions - Include only each player's nameand their NFLId
player_names <- players |> 
  select(nflId, displayName)

# Adds name of player and their position to plays we are predicting 
predicting_plays <- inner_join(predicting_plays, player_names)
```

```{r}
predicting_plays <- predicting_plays |>
  mutate(motioned = ifelse(!is.na(inMotionAtBallSnap), "Yes", "No")) # NAs represent players that never motioned at all
```

```{r}
# plays of players where they were set at the line
predicting_plays <- predicting_plays |> 
  mutate(set_on_line = ifelse(!is.na(motionSinceLineset), "Yes", "No"))
```

# Write CSVs -------------------------------------------

```{r}
#write.csv(predicting_plays, "offense_prediction_plays.csv")
```
