---
title: "Data Cleaning 2"
author: "Belle Schmidt"
date: "2025-01-02"
output: html_document
---

# Set Up ---------------------

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
library(stringr)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_8.csv")
tracking_week9 <- read.csv("tracking_week_9.csv")
```

# Only including players on offense ------------------------------------------

```{r}
possessionTeam <- plays |> 
  select(gameId, playId, possessionTeam)

teamName <- player_play |> 
  select(gameId, playId, nflId, teamAbbr, routeRan, motionSinceLineset, inMotionAtBallSnap, shiftSinceLineset)

players_on_offense <- teamName |> 
  inner_join(possessionTeam) |> 
  filter(teamAbbr == possessionTeam) |> 
  select(-teamAbbr, -possessionTeam)
```

```{r}
tracking_week1 <- inner_join(tracking_week1, players_on_offense) 

tracking_week2 <- inner_join(tracking_week2, players_on_offense)

tracking_week3 <- inner_join(tracking_week3, players_on_offense)

tracking_week4 <- inner_join(tracking_week4, players_on_offense)

tracking_week5 <- inner_join(tracking_week5, players_on_offense)

tracking_week6 <- inner_join(tracking_week6, players_on_offense)

tracking_week7 <- inner_join(tracking_week7, players_on_offense)

tracking_week8 <- inner_join(tracking_week8, players_on_offense)

tracking_week9 <- inner_join(tracking_week9, players_on_offense)
```

# Excluding Certain Plays -------------------

## Removing QB Spikes
```{r}
# Plays where the ball is not spiked 
non_spikes <- plays |> 
  filter(qbSpike == FALSE | is.na(qbSpike)) |> 
  select(gameId, playId)

tracking_week1 <- inner_join(tracking_week1, non_spikes)

tracking_week2 <- inner_join(tracking_week2, non_spikes)

tracking_week3 <- inner_join(tracking_week3, non_spikes)

tracking_week4 <- inner_join(tracking_week4, non_spikes)

tracking_week5 <- inner_join(tracking_week5, non_spikes)

tracking_week6 <- inner_join(tracking_week6, non_spikes)

tracking_week7 <- inner_join(tracking_week7, non_spikes)

tracking_week8 <- inner_join(tracking_week8, non_spikes)

tracking_week9 <- inner_join(tracking_week9, non_spikes)
```

## Removing QB Kneel Plays

```{r}
kneels <- plays |> 
  filter(qbKneel == 1) |> 
  select(gameId, playId)

tracking_week1 <- anti_join(tracking_week1, kneels)

tracking_week2 <- anti_join(tracking_week2, kneels)

tracking_week3 <- anti_join(tracking_week3, kneels)

tracking_week4 <- anti_join(tracking_week4, kneels)

tracking_week5 <- anti_join(tracking_week5, kneels)

tracking_week6 <- anti_join(tracking_week6, kneels)

tracking_week7 <- anti_join(tracking_week7, kneels)

tracking_week8 <- anti_join(tracking_week8, kneels)

tracking_week9 <- anti_join(tracking_week9, kneels)
```

## Removing QB Sneaks
```{r}
# remove QB sneaks from the data. This is okay to remove because QB sneaks are very predictable plays so the defense already knows the receiver isn't running a route. - There are still 40,438 receiver plays
non_sneaks <- plays |> 
  filter(is.na(qbSneak) | qbSneak == FALSE) |> 
  select(gameId, playId)

tracking_week1 <- inner_join(tracking_week1, non_sneaks)

tracking_week2 <- inner_join(tracking_week2, non_sneaks)

tracking_week3 <- inner_join(tracking_week3, non_sneaks)

tracking_week4 <- inner_join(tracking_week4, non_sneaks)

tracking_week5 <- inner_join(tracking_week5, non_sneaks)

tracking_week6 <- inner_join(tracking_week6, non_sneaks)

tracking_week7 <- inner_join(tracking_week7, non_sneaks)

tracking_week8 <- inner_join(tracking_week8, non_sneaks)

tracking_week9 <- inner_join(tracking_week9, non_sneaks)
```

# Dealing with play where two frames have "line_set" moment  ------------------------

```{r}
# Only occurred on one play in week 1
plays_with_new_set_line_week1 <- tracking_week1 |> 
  group_by(gameId, playId, nflId) |> 
  filter(event == "line_set") |> 
  summarise(num_times_set = n()) |> 
  ungroup() |> 
  filter(num_times_set == 2) |> 
  distinct(gameId, playId)

set_line_change_week1 <- tracking_week1 |> 
  inner_join(plays_with_new_set_line_week1) |> 
  distinct(gameId, playId)

set_line_change_week1

tracking_week1 <- tracking_week1 |>
  mutate(event = ifelse(gameId == 2022091103 & playId == 4434 & frameId == 4, "second_line_set", event))
```

# Only including players that are elligible receivers -----------------------------------

## Setting up the data to determine which players are eligible receivers 

### Adding variables indicating how many receivers lined up on the each side of the line, how many total receivers, and how far the line of scrimmage is from the end zone they are trying to score in 
```{r}
# add the absolute yard line (ranging 0 to 100),  number of receivers, and number of receivers on the left and right side variables to the tracking data (created receiver variables)
important_variables <- plays |> 
  mutate(num_receivers_left_temp = as.numeric(stringr::str_extract(receiverAlignment, "\\d{1}")),
         num_receivers_right_temp = as.numeric(stringr::str_extract(receiverAlignment, "(?<=x)[0-9]+")),
         total_receivers = num_receivers_left_temp + num_receivers_right_temp,
         absoluteYardLine = ifelse(possessionTeam != yardlineSide, yardlineNumber, 100 - yardlineNumber),  # makes it so the yard line variable is how far they are from their the end zone they are scoring in 
         absoluteYardLine = ifelse(yardlineNumber == 50, 50, absoluteYardLine),
         absoluteYardlineNumber = absoluteYardlineNumber - 10,
         num_receivers_left = ifelse(absoluteYardlineNumber == absoluteYardLine, num_receivers_left_temp, num_receivers_right_temp),
         num_receivers_right = ifelse(absoluteYardlineNumber == absoluteYardLine, num_receivers_right_temp, num_receivers_left_temp)) |> 
  select(gameId, playId, absoluteYardLine, total_receivers, num_receivers_left, num_receivers_right)

tracking_week1 <- inner_join(tracking_week1, important_variables)

tracking_week2 <- inner_join(tracking_week2, important_variables)

tracking_week3 <- inner_join(tracking_week3, important_variables)

tracking_week4 <- inner_join(tracking_week4, important_variables)

tracking_week5 <- inner_join(tracking_week5, important_variables)

tracking_week6 <- inner_join(tracking_week6, important_variables)

tracking_week7 <- inner_join(tracking_week7, important_variables)

tracking_week8 <- inner_join(tracking_week8, important_variables)

tracking_week9 <- inner_join(tracking_week9, important_variables)

```

### Creating a variable that indicates how far they are from the end zone they are trying to score in at the snap (rounded and not) 
```{r}

# Figuring out what yard line people are on at the snap

tracking_week1 <- tracking_week1 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine))  # all of these were off by less than a yard

tracking_week2 <- tracking_week2 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine))  # all of these were off by less than a yard

tracking_week3 <- tracking_week3 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine)) #  off by less than a yard or all eligible receivers already known 

tracking_week4 <- tracking_week4 |> 
   mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine)) # all of these were off by less than a yard

tracking_week5 <- tracking_week5 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine))

tracking_week6 <- tracking_week6 |> 
   mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine))

tracking_week7 <- tracking_week7 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine)) #  off by less than a yard

tracking_week8 <- tracking_week8 |> 
  mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine)) #  off by less than a yard

tracking_week9 <- tracking_week9 |> 
   mutate(yardLine_rounded = ifelse(playDirection == "left" & frameType == "SNAP", round(x - 10), NA),
         yardLine = ifelse(playDirection == "left" & frameType == "SNAP", x - 10, NA),
         yardLine_rounded = ifelse(playDirection == "right" & frameType == "SNAP", 100 - round(x - 10), yardLine), 
         yardLine = ifelse(playDirection == "right" & frameType == "SNAP", 100 - (x - 10), yardLine),
         yardLine_rounded= ifelse(frameType == "SNAP" & absoluteYardLine > yardLine, absoluteYardLine, yardLine)) #  off by less than a yard
```

### Indicating which players for sure were receivers on the play
```{r}
# identifying all players that ran routes as receivers

tracking_week1 <- tracking_week1 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA)) 

tracking_week2 <- tracking_week2 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week3 <- tracking_week3 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week4 <- tracking_week4 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week5 <- tracking_week5 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week6 <- tracking_week6 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week7 <- tracking_week7 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week8 <- tracking_week8 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week9 <- tracking_week9 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
```


### Subset Datasets (Plays with right number of receivers, not enough receivers, and too many receivers)
```{r}
# making a subset of the data of all plays where all receivers are already identified (finished), all plays where there are not enough receivers identified (not_enough_wr), and all plays where there are too many identified receivers (too_many_wr)

# week 1
finished_plays_week1 <- tracking_week1 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week1 <- tracking_week1 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week1 <- tracking_week1 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week1_finished <- inner_join(tracking_week1, finished_plays_week1) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week1_not_enough_wr <- inner_join(tracking_week1, not_enough_wr_plays_week1)

tracking_week1_too_many_wr <- inner_join(tracking_week1, too_many_wr_plays_week1)

# week 2
finished_plays_week2 <- tracking_week2 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week2 <- tracking_week2 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week2 <- tracking_week2 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week2_finished <- inner_join(tracking_week2, finished_plays_week2) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week2_not_enough_wr <- inner_join(tracking_week2, not_enough_wr_plays_week2)

tracking_week2_too_many_wr <- inner_join(tracking_week2, too_many_wr_plays_week2)

# week 3

finished_plays_week3 <- tracking_week3 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week3 <- tracking_week3 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week3 <- tracking_week3 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week3_finished <- inner_join(tracking_week3, finished_plays_week3) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week3_not_enough_wr <- inner_join(tracking_week3, not_enough_wr_plays_week3)

tracking_week3_too_many_wr <- inner_join(tracking_week3, too_many_wr_plays_week3)

# week 4
finished_plays_week4 <- tracking_week4 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week4 <- tracking_week4 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week4 <- tracking_week4 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week4_finished <- inner_join(tracking_week4, finished_plays_week4) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week4_not_enough_wr <- inner_join(tracking_week4, not_enough_wr_plays_week4)

tracking_week4_too_many_wr <- inner_join(tracking_week4, too_many_wr_plays_week4)

# week 5
finished_plays_week5 <- tracking_week5 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week5 <- tracking_week5 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week5 <- tracking_week5 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week5_finished <- inner_join(tracking_week5, finished_plays_week5) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week5_not_enough_wr <- inner_join(tracking_week5, not_enough_wr_plays_week5)

tracking_week5_too_many_wr <- inner_join(tracking_week5, too_many_wr_plays_week5)

# week 6
finished_plays_week6 <- tracking_week6 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week6 <- tracking_week6 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


too_many_wr_plays_week6 <- tracking_week6 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week6_finished <- inner_join(tracking_week6, finished_plays_week6) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week6_not_enough_wr <- inner_join(tracking_week6, not_enough_wr_plays_week6)

tracking_week6_too_many_wr <- inner_join(tracking_week6, too_many_wr_plays_week6)

# week 7

finished_plays_week7 <- tracking_week7 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week7 <- tracking_week7 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week7 <- tracking_week7 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week7_finished <- inner_join(tracking_week7, finished_plays_week7) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week7_not_enough_wr <- inner_join(tracking_week7, not_enough_wr_plays_week7)

tracking_week7_too_many_wr <- inner_join(tracking_week7, too_many_wr_plays_week7)


# week 8

finished_plays_week8 <- tracking_week8 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week8 <- tracking_week8 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week8 <- tracking_week8 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week8_finished <- inner_join(tracking_week8, finished_plays_week8) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week8_not_enough_wr <- inner_join(tracking_week8, not_enough_wr_plays_week8)

tracking_week8_too_many_wr <- inner_join(tracking_week8, too_many_wr_plays_week8)

# week 9

finished_plays_week9 <- tracking_week9 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers == num_receivers_identified) |> 
  select(gameId, playId)

not_enough_wr_plays_week9 <- tracking_week9 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)

too_many_wr_plays_week9 <- tracking_week9 |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers < num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified)


tracking_week9_finished <- inner_join(tracking_week9, finished_plays_week9) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week9_not_enough_wr <- inner_join(tracking_week9, not_enough_wr_plays_week9)

tracking_week9_too_many_wr <- inner_join(tracking_week9, too_many_wr_plays_week9)
```

### Identifying recievers on plays where the number of players that ran routes does not match receiver allignment ---------------------------

```{r}
# all plays where the not only receivers on the line were identified/ not enough wr were identified

not_enough_wr_plays <- rbind(tracking_week1_not_enough_wr, tracking_week2_not_enough_wr, tracking_week3_not_enough_wr, tracking_week4_not_enough_wr, tracking_week5_not_enough_wr, tracking_week6_not_enough_wr, tracking_week7_not_enough_wr, tracking_week8_not_enough_wr, tracking_week9_not_enough_wr) 

too_many_wr_plays <- rbind(tracking_week1_too_many_wr, tracking_week2_too_many_wr, tracking_week3_too_many_wr, tracking_week4_too_many_wr, tracking_week5_too_many_wr, tracking_week6_too_many_wr, tracking_week7_too_many_wr, tracking_week8_too_many_wr, tracking_week9_too_many_wr) 
```

#### Plays with one extra receiver ------------------------

```{r}
# identifies which plays only had one extra receiver and who the possible receivers are
one_extra_wr_plays <- too_many_wr_plays |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  mutate(dist_from_line = abs(absoluteYardLine - yardLine),
         diff_in_receivers = num_receivers_identified - total_receivers) |> 
  filter(diff_in_receivers == 1)

# adds position variable
positions <- players |> 
  select(nflId, position)

one_extra_wr_plays <- inner_join(one_extra_wr_plays, positions)

# identifies how many total plays have one extra receiver - 4,701
one_extra_wr_plays |> 
  distinct(gameId, playId)

# identify how many rb ran routes with plays with one extra receiver
num_rb_on_play <- one_extra_wr_plays |> 
  filter(position == "RB")|> 
  count(gameId, playId) |> 
  rename(num_rb = n) 
```

##### Plays with one running back 

```{r}
# only include plays with 1 running back on play - 4,473 
one_rb_plays <- num_rb_on_play |> 
  filter(num_rb == 1) |> 
  select(-num_rb)

finished_one_rb_plays <- one_extra_wr_plays |> 
  inner_join(one_rb_plays) |> 
  mutate(receiver_in_alignment = ifelse(position == "RB", FALSE, TRUE))
```

##### Plays with two running backs 

```{r}
# fixing plays with one extra receiver and two RB

two_rb_plays <- num_rb_on_play |> 
  filter(num_rb == 2) |> 
  select(-num_rb)

fixing_plays <- one_extra_wr_plays |> 
  inner_join(two_rb_plays)

# rb who were not receivers 
non_receivers_part1 <- fixing_plays |> 
  filter(num_receivers_left == 1) |> 
  arrange(y) |> 
  group_by(gameId, playId) |> 
  slice(2) |> 
  filter(position == "RB")

non_receivers_part2 <- fixing_plays |> 
  filter(num_receivers_left == 2) |> 
  arrange(y) |> 
  group_by(gameId, playId) |> 
  slice(3) |> 
  filter(position == "RB")

non_receivers_part3 <- fixing_plays |> 
  filter(num_receivers_left == 3) |> 
  arrange(y) |> 
  group_by(gameId, playId) |> 
  slice(4) |> 
  filter(position == "RB")

non_receivers <- rbind(non_receivers_part1, non_receivers_part2, non_receivers_part3) |> 
  mutate(receiver_in_alignment = FALSE)

fixed_play_num <- non_receivers |> 
  distinct(gameId, playId)

fixed_plays <- fixing_plays |> 
  inner_join(fixed_play_num)


 
fixing_plays |> 
  filter(num_receivers_left == 1) |> 
  arrange(gameId, playId, y)

motion_plays_w_two_rb <- fixing_plays |> 
  filter(motionSinceLineset == TRUE) |> 
  distinct(gameId, playId)
  
two_rb_no_motion <- anti_join(fixing_plays, motion_plays_w_two_rb)
  
rb_right_most <- two_rb_no_motion|> 
  group_by(gameId, playId) |> 
  slice_max(y, n = 1) |> 
  filter(position == "RB") |> 
  distinct(gameId, playId)

fixing_plays |> 
  inner_join(rb_right_most) |> 
  select(gameId, playId, y, position, dist_from_line, num_receivers_right, total_receivers, inMotionAtBallSnap, motionSinceLineset) |> 
  arrange(gameId, playId, y) 
```


```{r}

fixing_plays <- fixing_plays |> 
  group_by(gameId, playId, position) |> 
  mutate(obs_index = row_number()) |> 
  ungroup() |> 
  mutate(receiver_in_alignment = ifelse(position == "RB", NA, TRUE))

identifying_rb_wr <- fixing_plays |> 
  select(gameId, playId, position, dist_from_line, obs_index, y) |> 
  pivot_wider(
    names_from = c(position, obs_index), 
    values_from = c(dist_from_line, y),  
    names_sep = "_")
 

identifying_rb_wr

identifying_rb1_wr <- identifying_rb_wr|> 
  mutate(diff_rb = dist_from_line_RB_1 - dist_from_line_RB_2) |> 
  filter(diff_rb < -2) 

identifying_rb1_wr


fixed_plays_part1 <- fixing_plays |> 
  inner_join(identifying_rb1_wr) |> 
  mutate(receiver_in_alignment = ifelse(position == "RB" & obs_index == 1, TRUE, receiver_in_alignment))

identifying_rb2_wr <- identifying_rb_wr|> 
  mutate(diff_rb = dist_from_line_RB_1 - dist_from_line_RB_2) |> 
  filter(diff_rb > 2) |> 
  distinct(gameId, playId)


  
  
fixing_plays |> 
  inner_join(adding_number) |> 
  select(gameId, playId, y, position, )
  
  
  
identifying_rb_wr <- pivot_wider(names_from = position, values_from = dist_from_line)
  group_by(gameId, playId, ) |> 
  summarise(mean_dist = mean(dist_from_line))

|> 
  mutat
  distinct(gameId, playId)
  
  
  
  mutate(receiver_in_alignment = ifelse(position != "RB", TRUE, NA)) |> 
  
  


#|> 
  mutate(receiver_in_alignment = ifelse(position == "RB", FALSE, TRUE))
```



```{r}
careful_plays <- plays |> 
  filter(str_detect(tolower(playDescription), " eligible"))

players_first_last <- players|> 
  select(nflId, displayName) |> 
  mutate(firstName = str_extract(displayName, "^\\S+"),
         lastName = str_extract(displayName, "\\w+$")) 

players_abbreviated_name <- players_first_last |> 
  mutate(abbreviatedName = ifelse(displayName == "Deebo Samuel", "Deebo", NA),
         abbreviatedName = ifelse(displayName == "Kyler Murray", displayName, abbreviatedName),
         abbreviatedName = ifelse(displayName == "A.J. Brown", displayName, abbreviatedName))

duplicate_last_name <- players_abbreviated_name |> 
  filter(is.na(abbreviatedName)) |> 
  count(lastName) |> 
  filter(n > 1) |> 
  select(lastName)

duplicate_last_name

unique_players <- players_abbreviated_name |> 
  anti_join(duplicate_last_name) |> 
  mutate(abbreviatedFirstName = paste0(str_extract(firstName, "^."), "."),
         abbreviatedName = ifelse(is.na(abbreviatedName), paste(abbreviatedFirstName, lastName), abbreviatedName)) |> 
  select(nflId:displayName, abbreviatedName)

players_first_last |> 
  inner_join(duplicate_last_name)

```


# Figuring out which plays are run plays with no run pass option

```{r}
run_plays_week1 <- tracking_week1 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week2 <- tracking_week2 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week3 <- tracking_week3 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week4 <- tracking_week4 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week5 <- tracking_week5 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week6 <- tracking_week6 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week7 <- tracking_week7 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week8 <- tracking_week8 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week9 <- tracking_week9 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId)

run_plays <- rbind(run_plays_week1, run_plays_week2, run_plays_week3, run_plays_week4, run_plays_week5, run_plays_week6, run_plays_week7, run_plays_week8, run_plays_week9)

run_plays
```



```{r}

 <- 

plays_of_WR <- plays_of_WR |> 
  mutate(routeRan = ifelse(gameId %in% run_plays$gameId & playId %in% run_plays$playId & is.na(routeRan), "RUN BLOCK", routeRan))
```

```{r}
mis_match <- plays_of_WR |> 
  mutate(num_receivers_left = as.numeric(stringr::str_extract(receiverAlignment, "\\d{1}")),
         num_receivers_right = as.numeric(stringr::str_extract(receiverAlignment, "(?<=x)[0-9]+")),
         total_recievers = num_receivers_left + num_receivers_right) |> 
  count(gameId, playId, total_recievers) |> 
  filter(n > total_recievers)

mis_match


plays_of_WR |> 
  filter(gameId %in% mis_match$gameId & playId %in% mis_match$playId)
```

### Replacing NAs with saying no route ran 

```{r}
plays_of_WR <- plays_of_WR |> 
  mutate(routeRan = ifelse(pff_runPassOption == 1 & is.na(routeRan), "NO ROUTE", routeRan))
```

