---
title: "Other_Data_Cleaning_Ideas"
output: html_document
---

# NOTE: Some of these ideas were used on the previous data cleaning and some are still applicable now once I identify the receivers in alignment

##### Plays with two running backs 
```{r}
# fixing plays with one extra receiver and two RB

two_rb_plays <- num_rb_on_play |> 
  filter(num_rb == 2) |> 
  select(-num_rb)
 
fixing_plays <- one_extra_wr_plays |> 
  inner_join(two_rb_plays)
 
# rb who were not receivers 

non_receivers_part1 <- fixing_plays |> 
  filter(num_receivers_left == 1) |> 
  arrange(y) |> 
  group_by(gameId, playId) |> 
  slice(2) |> 
  filter(position != "RB")

non_receivers_part1

testing <- non_receivers_part1 |> 
  distinct(gameId, playId)

non_receivers_part1 |> 
  mutate(receiver_in_alignment = FALSE) |> 
  select(-dist_from_line, -diff_in_receivers, -position, -num_receivers_identified)

non_receivers_part2 <- fixing_plays |> 
  filter(num_receivers_left == 2) |> 
  arrange(y) |> 
  group_by(gameId, playId) |> 
  slice(3) |> 
  filter(position == "RB")

non_receivers_part3 <- fixing_plays |> 
  filter(num_receivers_left == 3) |> 
  arrange(y) |>  
  group_by(gameId, playId) |> 
  slice(4) |> 
  filter(position == "RB")

non_receivers <- rbind(non_receivers_part1, non_receivers_part2, non_receivers_part3) |> 
  mutate(receiver_in_alignment = FALSE)
 
fixed_play_num <- non_receivers |> 
  distinct(gameId, playId)

 
fixed_plays <- fixing_plays |> 
  inner_join(fixed_play_num)

 
fixing_plays |> 
  filter(num_receivers_left == 1) |> 
  arrange(gameId, playId, y)

motion_plays_w_two_rb <- fixing_plays |> 
  filter(motionSinceLineset == TRUE) |> 
  distinct(gameId, playId)
 
two_rb_no_motion <- anti_join(fixing_plays, motion_plays_w_two_rb)
 
rb_right_most <- two_rb_no_motion|> 
  group_by(gameId, playId) |> 
  slice_max(y, n = 1) |> 
  filter(position == "RB") |> 
  distinct(gameId, playId)

fixing_plays |> 
  inner_join(rb_right_most) |> 
  select(gameId, playId, y, position, dist_from_line, num_receivers_right, total_receivers, inMotionAtBallSnap, motionSinceLineset) |> 
  arrange(gameId, playId, y) 

fixing_plays |> 
  inner_join(testing) |> 
  arrange(gameId, playId, y)
```

```{r}
plays |> 
  inner_join(testing) |> 
  arrange(gameId, playId)
```
 
```{r}
fixing_plays <- fixing_plays |> 
  group_by(gameId, playId, position) |> 
  mutate(obs_index = row_number()) |> 
  ungroup() |> 
  mutate(receiver_in_alignment = ifelse(position == "RB", NA, TRUE))

identifying_rb_wr <- fixing_plays |> 
  select(gameId, playId, position, dist_from_line, obs_index, y) |> 
  pivot_wider(
  names_from = c(position, obs_index), 
  values_from = c(dist_from_line, y),  
  names_sep = "_")

identifying_rb_wr

identifying_rb1_wr <- identifying_rb_wr|> 
  mutate(diff_rb = dist_from_line_RB_1 - dist_from_line_RB_2) |> 
  filter(diff_rb < -2) 

identifying_rb1_wr

fixed_plays_part1 <- fixing_plays |> 
  inner_join(identifying_rb1_wr) |> 
  mutate(receiver_in_alignment = ifelse(position == "RB" & obs_index == 1, TRUE, receiver_in_alignment))

identifying_rb2_wr <- identifying_rb_wr|> 
  mutate(diff_rb = dist_from_line_RB_1 - dist_from_line_RB_2) |> 
  filter(diff_rb > 2) |> 
  distinct(gameId, playId)
 
fixing_plays |> 
  inner_join(adding_number) |> 
  select(gameId, playId, y, position, )

identifying_rb_wr <- pivot_wider(names_from = position, values_from = dist_from_line)
  group_by(gameId, playId, ) |> 
  summarise(mean_dist = mean(dist_from_line))
```
 
```{r}
careful_plays <- plays |> 
  filter(str_detect(tolower(playDescription), " eligible"))

players_first_last <- players|> 
  select(nflId, displayName) |> 
  mutate(firstName = str_extract(displayName, "^\\S+"),
          lastName = str_extract(displayName, "\\w+$")) 
 
players_abbreviated_name <- players_first_last |> 
  mutate(abbreviatedName = ifelse(displayName == "Deebo Samuel", "Deebo", NA),
         abbreviatedName = ifelse(displayName == "Kyler Murray", displayName, abbreviatedName),
         abbreviatedName = ifelse(displayName == "A.J. Brown", displayName, abbreviatedName))

duplicate_last_name <- players_abbreviated_name |> 
  filter(is.na(abbreviatedName)) |> 
  count(lastName) |> 
  filter(n > 1) |> 
  select(lastName)

duplicate_last_name

unique_players <- players_abbreviated_name |> 
  anti_join(duplicate_last_name) |> 
  mutate(abbreviatedFirstName = paste0(str_extract(firstName, "^."), "."),
         abbreviatedName = ifelse(is.na(abbreviatedName), paste(abbreviatedFirstName, lastName), abbreviatedName)) |> 
  select(nflId:displayName, abbreviatedName)
 
players_first_last |> 
  inner_join(duplicate_last_name)

```
 
# Figuring out which plays are run plays with no run pass option
 
```{r}
run_plays_week1 <- tracking_week1 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week2 <- tracking_week2 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week3 <- tracking_week3 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week4 <- tracking_week4 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week5 <- tracking_week5 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week6 <- tracking_week6 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week7 <- tracking_week7 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 
 
run_plays_week8 <- tracking_week8 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId) 

run_plays_week9 <- tracking_week9 |> 
  filter(event == "handoff" | event == "run") |> 
  distinct(gameId, playId)
 
run_plays <- rbind(run_plays_week1, run_plays_week2, run_plays_week3, run_plays_week4, run_plays_week5, run_plays_week6, run_plays_week7, run_plays_week8, run_plays_week9)
 
run_plays

```
 
```{r}
plays_of_WR <- plays_of_WR |> 
  mutate(routeRan = ifelse(gameId %in% run_plays$gameId & playId %in% run_plays$playId & is.na(routeRan), "RUN BLOCK", routeRan))

```

```{r}
mis_match <- plays_of_WR |> 
  mutate(num_receivers_left = as.numeric(stringr::str_extract(receiverAlignment, "\\d{1}")),
         num_receivers_right = as.numeric(stringr::str_extract(receiverAlignment, "(?<=x)[0-9]+")),
         total_recievers = num_receivers_left + num_receivers_right) |> 
  count(gameId, playId, total_recievers) |> 
  filter(n > total_recievers)

mis_match
 
plays_of_WR |> 
  filter(gameId %in% mis_match$gameId & playId %in% mis_match$playId)
```

### Replacing NAs with saying no route ran 
 
```{r}
plays_of_WR <- plays_of_WR |> 
  mutate(routeRan = ifelse(pff_runPassOption == 1 & is.na(routeRan), "NO ROUTE", routeRan))
```
 
### Indicating which players for sure were receivers on the play
```{r}
# identifying all players that ran routes as receivers
tracking_week1 <- tracking_week1 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA)) 
 
tracking_week2 <- tracking_week2 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
 
tracking_week3 <- tracking_week3 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))

tracking_week4 <- tracking_week4 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
 
tracking_week5 <- tracking_week5 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
  
tracking_week6 <- tracking_week6 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
 
tracking_week7 <- tracking_week7 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
 
tracking_week8 <- tracking_week8 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
 
tracking_week9 <- tracking_week9 |>
  mutate(receiver = ifelse(!is.na(routeRan), TRUE, NA))
```

### Subset Datasets (Plays with right number of receivers, not enough receivers, and too many receivers)
```{r}

# making a subset of the data of all plays where all receivers are already identified (finished), all plays where there are not enough receivers identified (not_enough_wr), and all plays where there are too many identified receivers (too_many_wr)

finished_plays <- function(week_num){
week_num |> 
    filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
    count(gameId, playId, total_receivers_in_alignment) |> 
    rename(num_receivers_identified = n) |> 
    filter(total_receivers_in_alignment == num_receivers_identified) |> 
    select(gameId, playId)
}
 
finished_plays_week1 <- finished_plays(tracking_week1)

finished_plays_week2 <- finished_plays(tracking_week2)

finished_plays_week3 <- finished_plays(tracking_week3)

finished_plays_week4 <- finished_plays(tracking_week4)

finished_plays_week5 <- finished_plays(tracking_week5)

finished_plays_week6 <- finished_plays(tracking_week6)

finished_plays_week7 <- finished_plays(tracking_week7)

finished_plays_week8 <- finished_plays(tracking_week8)

finished_plays_week9 <- finished_plays(tracking_week9)

 
tracking_week1_finished <- inner_join(tracking_week1, finished_plays_week1) |> 
  filter(frameType != "AFTER_SNAP") 
 
tracking_week2_finished <- inner_join(tracking_week2, finished_plays_week2) |> 
  filter(frameType != "AFTER_SNAP") 
 
tracking_week3_finished <- inner_join(tracking_week3, finished_plays_week3) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week4_finished <- inner_join(tracking_week4, finished_plays_week4) |> 
  filter(frameType != "AFTER_SNAP") 
  
tracking_week5_finished <- inner_join(tracking_week5, finished_plays_week5) |> 
  filter(frameType != "AFTER_SNAP") 
 
tracking_week6_finished <- inner_join(tracking_week6, finished_plays_week6) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week7_finished <- inner_join(tracking_week7, finished_plays_week7) |> 
  filter(frameType != "AFTER_SNAP") 

tracking_week8_finished <- inner_join(tracking_week8, finished_plays_week8) |> 
  filter(frameType != "AFTER_SNAP") 
 
tracking_week9_finished <- inner_join(tracking_week9, finished_plays_week9) |> 
  filter(frameType != "AFTER_SNAP") 
 
```
 
```{r}
not_enough_wr_plays <- function(week_num) {
  week_num |> 
   filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  count(gameId, playId, total_receivers_in_alignment) |> 
  rename(num_receivers_identified = n) |> 
  filter(total_receivers_in_alignment > num_receivers_identified) |> 
  select(gameId, playId, num_receivers_identified) 
}

not_enough_wr_plays_week1 <- not_enough_wr_plays(tracking_week1)

not_enough_wr_plays_week2 <- not_enough_wr_plays(tracking_week2)

not_enough_wr_plays_week3 <- not_enough_wr_plays(tracking_week3)

not_enough_wr_plays_week4 <- not_enough_wr_plays(tracking_week4)

not_enough_wr_plays_week5 <- not_enough_wr_plays(tracking_week5)

not_enough_wr_plays_week6 <- not_enough_wr_plays(tracking_week6)

not_enough_wr_plays_week7 <- not_enough_wr_plays(tracking_week7)

not_enough_wr_plays_week8 <- not_enough_wr_plays(tracking_week8)

not_enough_wr_plays_week9 <- not_enough_wr_plays(tracking_week9)


tracking_week1_not_enough_wr <- inner_join(tracking_week1, not_enough_wr_plays_week1)

tracking_week2_not_enough_wr <- inner_join(tracking_week2, not_enough_wr_plays_week2)

tracking_week3_not_enough_wr <- inner_join(tracking_week3, not_enough_wr_plays_week3)

tracking_week4_not_enough_wr <- inner_join(tracking_week4, not_enough_wr_plays_week4)

tracking_week5_not_enough_wr <- inner_join(tracking_week5, not_enough_wr_plays_week5)

tracking_week6_not_enough_wr <- inner_join(tracking_week6, not_enough_wr_plays_week6)

tracking_week7_not_enough_wr <- inner_join(tracking_week7, not_enough_wr_plays_week7)

tracking_week8_not_enough_wr <- inner_join(tracking_week8, not_enough_wr_plays_week8)

tracking_week9_not_enough_wr <- inner_join(tracking_week9, not_enough_wr_plays_week9)
 
```
 
```{r}
too_many_wr_plays <- function(week_num){
  week_num |> 
    filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
    count(gameId, playId, total_receivers_in_alignment) |> 
    rename(num_receivers_identified = n) |> 
    filter(total_receivers_in_alignment < num_receivers_identified) |> 
    select(gameId, playId, num_receivers_identified)
} 

too_many_wr_plays_week1 <- too_many_wr_plays(tracking_week1)

too_many_wr_plays_week2 <- too_many_wr_plays(tracking_week2)

too_many_wr_plays_week3 <- too_many_wr_plays(tracking_week3)

too_many_wr_plays_week4 <- too_many_wr_plays(tracking_week4)

too_many_wr_plays_week5 <- too_many_wr_plays(tracking_week5)

too_many_wr_plays_week6 <- too_many_wr_plays(tracking_week6)

too_many_wr_plays_week7 <- too_many_wr_plays(tracking_week7)

too_many_wr_plays_week8 <- too_many_wr_plays(tracking_week8)

too_many_wr_plays_week9 <- too_many_wr_plays(tracking_week9)

tracking_week1_too_many_wr <- inner_join(tracking_week1, too_many_wr_plays_week1)

tracking_week2_too_many_wr <- inner_join(tracking_week2, too_many_wr_plays_week2)

tracking_week3_too_many_wr <- inner_join(tracking_week3, too_many_wr_plays_week3)

tracking_week4_too_many_wr <- inner_join(tracking_week4, too_many_wr_plays_week4)

tracking_week5_too_many_wr <- inner_join(tracking_week5, too_many_wr_plays_week5)

tracking_week6_too_many_wr <- inner_join(tracking_week6, too_many_wr_plays_week6)

tracking_week7_too_many_wr <- inner_join(tracking_week7, too_many_wr_plays_week7)

tracking_week8_too_many_wr <- inner_join(tracking_week8, too_many_wr_plays_week8)

tracking_week9_too_many_wr <- inner_join(tracking_week9, too_many_wr_plays_week9)
```
 
```{r}
# combining all the various plays into three larger datasets 

# all plays where there were less players that ran routes than included in the receiver alignment 
not_enough_wr_plays <- rbind(tracking_week1_not_enough_wr, tracking_week2_not_enough_wr, tracking_week3_not_enough_wr, tracking_week4_not_enough_wr, tracking_week5_not_enough_wr, tracking_week6_not_enough_wr, tracking_week7_not_enough_wr, tracking_week8_not_enough_wr, tracking_week9_not_enough_wr) 

# 616 plays have less players running routes than receiver alignment 
not_enough_wr_plays |> 
  distinct(gameId, playId)
 
# all plays where more players ran routes than were included in the receiver alignment 
too_many_wr_plays <- rbind(tracking_week1_too_many_wr, tracking_week2_too_many_wr, tracking_week3_too_many_wr, tracking_week4_too_many_wr, tracking_week5_too_many_wr, tracking_week6_too_many_wr, tracking_week7_too_many_wr, tracking_week8_too_many_wr, tracking_week9_too_many_wr) 

# 4992 plays have more players running routes than on the receiver alignment 
too_many_wr_plays |> 
  distinct(gameId, playId)
 
# all plays where the number of receivers in the receiver alignment matches the number of players that ran routes
finished_plays <- rbind(tracking_week1_finished, tracking_week2_finished, tracking_week3_finished, tracking_week4_finished, tracking_week5_finished, tracking_week6_finished, tracking_week7_finished, tracking_week8_finished, tracking_week9_finished)

finished_plays <- finished_plays |> 
  mutate(receiver_in_alignment = ifelse(!is.na(routeRan), TRUE, FALSE))
  
# 3,040 plays have the same number of receivers running routes as in the receiver alignment
finished_plays |> 
  distinct(gameId, playId)
```
 
### Identifying recievers included in the alignment on plays where the number of players that ran routes does not match receiver allignment ----
 
#### Plays with one extra receiver ------------------------

```{r}
# identifies which plays only had one extra receiver and who the possible receivers arE
one_extra_wr_plays <- too_many_wr_plays |> 
  filter(frameType == "SNAP",
         !is.na(routeRan)) |> 
  mutate(dist_from_line = abs(yardsTillTouchdown - yardLine),
         diff_in_receivers = num_receivers_identified - total_receivers_in_alignment) |> 
  filter(diff_in_receivers == 1)
 
# adds position variable
positions <- players |> 
  select(nflId, position)
  
one_extra_wr_plays <- inner_join(one_extra_wr_plays, positions)
 
# identifies how many total plays have one extra receiver - 4,701
one_extra_wr_plays |> 
  distinct(gameId, playId)
  
# identify how many rb ran routes with plays with one extra receiver
num_rb_on_play <- one_extra_wr_plays |> 
  filter(position == "RB")|> 
  count(gameId, playId) |> 
  rename(num_rb = n) 
```
 
##### Plays with one running back 
 
```{r}
# only include plays with 1 running back on play - 4,473 

one_rb_plays <- num_rb_on_play |> 
  filter(num_rb == 1) |> 
  select(-num_rb)
  
# made it so that the rb is the player who is not included in the alignment
finished_one_rb_plays <- one_extra_wr_plays |> 
  inner_join(one_rb_plays) |> 
  mutate(receiver_in_alignment = ifelse(!is.na(routeRan) & position != "RB", TRUE, FALSE)) |> 
  select(-dist_from_line, -diff_in_receivers, -position, -num_receivers_identified)
```
 
# Plays where the receivers in alignment are identified (plays with one too many receivers and one rb and ones with correct number of receivers)

```{r}
finished_plays <- rbind(finished_plays, finished_one_rb_plays)
 
# Finished 7,513 plays
finished_plays |> 
  distinct(gameId, playId)

 
```

### Figuring out how many players are on the left and how many are on the right 

```{r}
520
 
players_predicting |> 
  filter(frameType == "SNAP") |> 
  count(gameId, playId, receiver_in_left_alignment) |> 
  mutate(num_left_receivers = ifelse(receiver_in_left_alignment == TRUE))
```
