---
title: "Play Animation"
author: "Belle Schmidt"
date: "2024-12-02"
output: ''
---

```{r}
# code to install ggfootbaall
#remotes::install_github("robkravec/ggfootball")
```


```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_9.csv")
```

```{r}
player_play |> 
  select(gameId, playId, nflId, motionSinceLineset, inMotionAtBallSnap, shiftSinceLineset)
```

```{r}
# Filter data for specific gameId, nflId, and displayName
filtered_data <- tracking_week1 |> 
  filter(gameId == 2022091200, nflId == 35459, displayName == "Kareem Jackson")
```

```{r}
# View the filtered data
head(filtered_data)
```

```{r}
# Sanders play
PHI_JAG_Game <- plays |> 
  filter(
    gameId==2022100209,
    quarter == 3,
    possessionTeam=="PHI",
    gameClock >= "03:35" & gameClock <= "03:45"  # Filter for gameClock between 3:35 and 3:45
  ) |> 
  arrange(gameClock)  

View(PHI_JAG_Game)
```

```{r}
# filter play by play data for a specific play. In this case, the Sander's play
filtered_data_sanders <- tracking_week4  |> 
  filter(
    gameId==2022100209,
    club == "PHI" | club == "JAX",
    grepl("2022-10-02", time),
    playId==2652
  ) |> 
arrange(desc(s)) 
View(filtered_data_sanders)
```

# start of code for play animation --------------------------------------------
# TO DO: make the animation prettier, 
# get a better understanding of arrow orientation/make it flow smoother

```{r}
# all players on the field nflId's during sandars run play-----------------
players_on_field <- filtered_data_sanders |> 
  select(nflId, club, jerseyNumber) |> 
  distinct()

View(players_on_field)
```

```{r}
# list of nflId's (players) on the field during Sander's run play
nfl_Ids <- c(37266, 39950, 43352, 43368, 43787, 44902, 44926, 
                46118, 46269, 47790, 47834, 47836, 52461, 52481, 
                52553, 52608, 53439, 53462, 53466, 53494, 54492, 54758)
```

```{r}
# filter tracking data to include specific players for this play and time range
filtered_sanders_run_play <- tracking_week4 |> 
  filter(
    gameId == 2022100209,  
    playId == 2652, 
    nflId %in% nfl_Ids,  
    grepl("2022-10-02", time),
    !is.na(x) & !is.na(y) & !is.na(o) 
  ) 
View(filtered_sanders_run_play)
```

```{r}
# Define team colors
team_colors <- c("PHI" = "#004C54", "JAG" = "#D7A22A")
```


# Create an animation of all players' movement during the run play (3rd Q, PHI vs. JAG) ----------------

```{r}
animation_all_players <- ggfootball(left_endzone = "#004C54", right_endzone = "#D7A22A", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = filtered_sanders_run_play, 
             aes(x = x, y = y, fill = club), size = 4, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add arrows for player orientation
  geom_segment(data = filtered_sanders_run_play, 
               aes(x = x, y = y, 
                   xend = pmin(pmax(x + cos(o * pi / 180) * 5, 0), 120),  # multiplier to adjust arrow length
                   yend = pmin(pmax(y + sin(o * pi / 180) * 5, 0), 53.3),   
                   color = club),
               arrow = arrow(type = "closed", length = unit(0.05, "inches")), size = 0.2, color = "black", show.legend = FALSE) +
  # add labels for jersey numbers
  geom_text(data = filtered_sanders_run_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  scale_color_manual(values = c("#004C54", "white")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#004C54", "white")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During Run Play", 
       subtitle = "Frame: {frame_time}", 
       x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```


# create the animation with transition by frameId
```{r}
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

```

# display animation with proper Frames Per Second (FPS)
```{r}
total_frames <- length(unique(filtered_sanders_run_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 25)
```

# create an animation of Miles Sanders run play (only the runner)--------------------
```{r}
# filter play by play data for a specific play (only include sanders)
filtered_sanders_coordinates <- tracking_week4  |> 
  filter(
    gameId==2022100209,
    club == "PHI" | club == "JAX",
    grepl("2022-10-02", time),
    playId==2652, 
    nflId =="47836"
  ) |> 
arrange(desc(s)) 

filtered_sanders_coordinates
```


```{r}
animation <- ggfootball(left_endzone = "#004C54", right_endzone = "#D7A22A", field_alpha = 0.7) +
  # Plot the player's position over time as points
  geom_point(data = filtered_sanders_coordinates, aes(x = x, y = y, color = s), size = 4) + 
  # Add arrows for player orientation
  geom_segment(data = filtered_sanders_coordinates, 
               aes(x = x, y = y, 
                   xend = x + cos(o * pi / 180) * 2,  # Adjust multiplier for arrow length
                   yend = y + sin(o * pi / 180) * 2),  # Adjust multiplier for arrow length
               arrow = arrow(type = "open", length = unit(0.1, "inches")), size = 0.5) +
  # Set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  labs(title = "Player Movement: Miles Sanders", x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


# Create the animation
```{r}
animation <- animation + 
  transition_time(frameId) +  # Animate based on frameId (assuming frames are sequential)
  ease_aes('linear')  # Smooth linear transition
```


# Display the animation with the correct frames per second
```{r}
animate(animation, nframes = nrow(filtered_sanders_coordinates), fps = 25)
```


#--------------------------

# Filter tracking data to only include the deep ball play I want to look at

```{r}
deep_pass_play <- tracking_week1 |> 
  filter(gameId == "2022091100", playId == "1260",
         club == "NO" | club == "ATL",
    grepl("2022-10-02", time),
  distinct(club)
```

# Create an animation of all players' movement during the run play (3rd Q, PHI vs. JAG) ----------------

```{r}
animation_all_players <- ggfootball(left_endzone = "#004C54", right_endzone = "#D7A22A", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = deep_pass_play, 
             aes(x = x, y = y, fill = club), size = 4, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add arrows for player orientation
  geom_segment(data = deep_pass_play, 
               aes(x = x, y = y, 
                   xend = pmin(pmax(x + cos(o * pi / 180) * 5, 0), 120),  # multiplier to adjust arrow length
                   yend = pmin(pmax(y + sin(o * pi / 180) * 5, 0), 53.3),   
                   color = club),
               arrow = arrow(type = "closed", length = unit(0.05, "inches")), size = 0.2, color = "black", show.legend = FALSE) +
  # add labels for jersey numbers
  geom_text(data = deep_pass_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  scale_color_manual(values = c("#004C54", "white")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#004C54", "white")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During Run Play", 
       subtitle = "Frame: {frame_time}", 
       x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```


# create the animation with transition by frameId
```{r}
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

```

# display animation with proper Frames Per Second (FPS)
```{r}
total_frames <- length(unique(deep_pass_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 25)
```

# create an animation of Miles Sanders run play (only the runner)--------------------
```{r}
# filter play by play data for a specific play (only include sanders)
filtered_sanders_coordinates <- tracking_week4  |> 
  filter(
    gameId==2022100209,
    club == "PHI" | club == "JAX",
    grepl("2022-10-02", time),
    playId==2652, 
    nflId =="47836"
  ) |> 
arrange(desc(s)) 

filtered_sanders_coordinates
```


```{r}
animation <- ggfootball(left_endzone = "#004C54", right_endzone = "#D7A22A", field_alpha = 0.7) +
  # Plot the player's position over time as points
  geom_point(data = filtered_sanders_coordinates, aes(x = x, y = y, color = s), size = 4) + 
  # Add arrows for player orientation
  geom_segment(data = filtered_sanders_coordinates, 
               aes(x = x, y = y, 
                   xend = x + cos(o * pi / 180) * 2,  # Adjust multiplier for arrow length
                   yend = y + sin(o * pi / 180) * 2),  # Adjust multiplier for arrow length
               arrow = arrow(type = "open", length = unit(0.1, "inches")), size = 0.5) +
  # Set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  labs(title = "Player Movement: Miles Sanders", x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```


# Create the animation
```{r}
animation <- animation + 
  transition_time(frameId) +  # Animate based on frameId (assuming frames are sequential)
  ease_aes('linear')  # Smooth linear transition
```


# Display the animation with the correct frames per second
```{r}
animate(animation, nframes = nrow(filtered_sanders_coordinates), fps = 25)
```

# Get unique club names
unique_clubs <- tracking_w_1 |> 
  select(club) |> 
  distinct()

# View the unique club names
print(unique_clubs)
View(unique_clubs)


# View the first few rows
head(data)

# other random EDA -------------------------------------------------
# nflreadr::load_data()

pbp <- load_participation(season = 2023, include_pbp = TRUE)
View(pbp)

pass_rush_epa <- pbp |>
  filter (pass == 1) |> 
  group_by(defteam) |> 
  summarise(plays=n(),
            avg_pass_rushers = round(mean(number_of_pass_rushers,na.rm=T),3),
            epa = mean (epa, na.rm=T))

ggplot2::ggplot(pass_rush_epa,aes(x=avg_pass_rushers, y=epa)) +
  geom_mean_lines(aes(x0=avg_pass_rushers, y0= epa)) +
  geom_nfl_logos(aes(team_abbr = defteam),width=0.055) + 
  theme_classic()+
  ggtitle("EPA/play on Average Amount of Pass Rushers",
          "2023 Season Through Week 12") +
  xlab('Average Amount of Pass Rushers') +
  ylab('EPA/play') +
  labs(caption = 'Data: nflreadr')

participation <- load_participation(seasons = 2023)
View(participation)

rosters_2023 <- load_rosters(seasons = 2023)
View(rosters_2023)

pbp_2023 <- load_pbp(seasons = 2023)
View(pbp_2023)

schedules_2023 <- load_schedules(seasons = 2023)
View(schedules_2023)

snap_counts_2023 <- load_snap_counts(seasons = 2023)
View(snap_counts)

pbp_2020 <- nflreadr::load_pbp(2020) |> 
  dplyr::filter(season_type == "REG") |>
  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))

offense <- pbp_2020  |>
  dplyr::group_by(team = posteam) |>
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))

defense <- pbp_2020  |>
  dplyr::group_by(team = defteam) |>
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))

combined <- offense |>
  dplyr::inner_join(defense, by = "team")

qbs <- pbp_2020  |>
  dplyr::filter(pass == 1 | rush == 1) |>
  dplyr::filter(down %in% 1:4) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    name = dplyr::first(name),
    team = dplyr::last(posteam),
    plays = dplyr::n(),
    qb_epa = mean(qb_epa, na.ram = TRUE)
  ) |>
  dplyr::filter(plays > 200) |>
  dplyr::slice_max(qb_epa, n = 10)
# ----------------
# Logos in Scatter Plots
ggplot2::ggplot(combined, aes(x = off_epa, y = def_epa)) +
  ggplot2::geom_abline(slope = -1.5, intercept = seq(0.4, -0.3, -0.1), alpha = .2) +
  nflplotR::geom_mean_lines(aes(x0 = off_epa , y0 = def_epa)) +
  nflplotR::geom_nfl_logos(aes(team_abbr = team), width = 0.065, alpha = 0.7) +
  ggplot2::labs(
    x = "Offense EPA/play",
    y = "Defense EPA/play",
    caption = "Data: @nflfastR",
    title = "2020 NFL Offensive and Defensive EPA per Play"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    plot.title.position = "plot",
    plot.background = ggplot2::element_rect(fill = "#F0F0F0")
  ) +
  ggplot2::scale_y_reverse()

