---
title: "Route Prediction"
author: "Belle Schmidt"
date: "2024-11-19"
output: html_document
---

# Set Up ---------------------

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_9.csv")
```

# Determining What Plays To Include in Route Running Predicition ------------------

## Setting up the data for more exploration
```{r}
# add play information to individual player play by play data
player_play_w_info <- player_play |> 
  inner_join(plays, join_by(gameId, playId))
```

## Decided to focus on only WR 
```{r}
# Players WR Positions - Include only position of each player, their name, and their NFLId
positions <- players |> 
  filter(position == "WR") |> 
  select(nflId, displayName, position)

# All play information of WRs that ran at least one route - there are 40,940
plays_of_route_runners <- inner_join(player_play_w_info, positions)
```

## Finding the distribution of the routes ran - it is a little worrisome that so many routes are NAs. There is also a very large disproportion each route is run
```{r}
# The number of times and the proportion of the time each route is run by WR or TEs 
plays_of_route_runners |> 
  group_by(routeRan) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```

# Figuring out what some of the NAs could mean -------------------

```{r}
# Plays where at least one receiver or tight end did not run a route - there are 6,874 plays
no_route_ran <- plays_of_route_runners |> 
  filter(is.na(routeRan)) |> 
  distinct(gameId, playId)
```

## Removing Deadplays
```{r}
# Plays where the ball is spiked or the QB kneels - there are 188
deadplays <- plays |> 
  filter(is.na(offenseFormation))

# removes kneeling and spike plays - now there are 6,690 plays
no_route_ran <- anti_join(no_route_ran, deadplays)


# removes kneeling and spike plays from WR play by play info - there are now 40,690 observations

plays_of_route_runners <- anti_join(plays_of_route_runners, deadplays, join_by(gameId, playId))
```

## Removing QB Sneaks
```{r}
# remove QB sneaks from the data. This is okay to remove because QB sneaks are very predictable plays so the defense already knows the receiver isn't running a route. - There are still 40,438 reciever plays
plays_of_route_runners <- plays_of_route_runners |> 
  filter(is.na(qbSneak) | qbSneak == FALSE)

# Creating a list of plays where qb sneaks are not included
valid_plays <- plays_of_route_runners |> 
  distinct(gameId, playId)

# Remove QB sneaks from the list of plays where at least one WR with an NA route - 6,571 plays
no_route_ran <- no_route_ran|> 
  inner_join(valid_plays)

```

## Replacing NAs with saying no route ran 
```{r}
plays_of_route_runners <- plays_of_route_runners |> 
  mutate(routeRan = replace_na(routeRan, "NO ROUTE"))
```

# Exploratory Data Analysis --------------------------------------------------------

## Distribution of Variables ------------------------------------

### Routes Ran -----------------

```{r}
# The number of times and the proportion of the time each route is run by WR (Table)
prop_route <- plays_of_route_runners |> 
  group_by(routeRan) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```


```{r, fig.cap= "There is a very large disproportion of the proportion of times each route is ran. Specifically a lot of the time receivers ran no route."}

# Plot of distribution of routes ran
prop_route |> 
  ggplot(aes(x = fct_reorder(routeRan, prop), y = prop, fill = routeRan)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Routes Ran By Wide Recievers",
    x = "Route Ran",
    y = "Proportion Of The Time Route Was Ran"
  )
```

#### Defining the Different Routes 
**Wheel:** When the reciever runs kind of an L shaped path. Out and then down field - most commonly run by RB

**Angle:** when the player cuts out 45 degrees and then takes a 90 degree turn and cuts back in (often RB, FB, or TE) - makes sense why few WR run it

**Screen:** Various types. More of a basic screen is where RB catches ball and lineman block. Then Jailbreak screen is where the receiver out wide is the target. Two receivers and a lineman are blockers. Bubble screen is where WR drops back slightly and two receivers block. Drop back screen looks like drop back pass but one WR runs route and other WR block. Could be TE too? 

**Flat:** when the receiver runs a quick short route straight then towards the sideline

**Corner:** when the receiver runs down the field and then turns outward 45degrees (opposite direction of post)

**Slant:** when the receiver runs shortly straight and then turns inward 45 degrees (like short version of post)

**Post:** when the receiver runs 10 to 20 yards down field and then turns 45 degrees and runs towards the goal posts 

**Out:** when the receiver runs down the field and then cuts out and runs parallel to the line of scrimmage

**In:** when the receiver runs down the field and then cuts in and runs parallel to the line of scrimmage

**Cross:** when the receiver crosses across the field

**Hitch:** when the receiver runs a set number of yards (normally 5 to 6) and then turns to face the QB 

**Go:** When the wide receiver runs straight down field as fast as they can

### Formations ------------------------

```{r}
# Table of the number of times and the proportion of the time each formation is used
prop_formation <- plays_of_route_runners |> 
  group_by(offenseFormation) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```


```{r}
# Graph showing distribution of formations 
prop_formation |> 
  ggplot(aes(x = fct_reorder(offenseFormation, prop), y = prop, fill = offenseFormation)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Formation Usage",
    x = "Formation",
    y = "Proportion of Time Formation Was Used"
  )
```

#### Defining the Different Formations

**Shotgun:** When the quarterback is several yards back from the center and has more time to read the defense. In this case, when the ball is snapped, the quarterback is not grabbing it directly from the center.

**Single back:** When the quarterback is under center (grabs the ball directly from the center) and there is a single RB about 5 yards back

**Empty:** When there are no backs in the backfield so the quarterback is the only player in the backfield

**Pistol:** When the quarterback is a couple yards behind the center (closer than for shotgun though) and there is a running back right behind him

**Wildcat:** when the ball is snapped to a position player other than the QB... might want to remove these

**Jumbo:** when an offensive lineup is designed for maximum blocking power (often close to goal line). Normally sub a receiver for an extra TE.


### Receiver Alignment ---------------------------

```{r}
# Table showing he number of times and the proportion of the time each receiver alignment is used
receiver_align_prop <- plays_of_route_runners |> 
  group_by(receiverAlignment) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```

```{r}
# Graph showing distribution of receiver alignment 

receiver_align_prop |> 
  ggplot(aes(x = fct_reorder(receiverAlignment, prop), y= prop, fill = receiverAlignment)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Receiver Alignment Usage",
    x = "Receiver Alignment",
    y = "Proportion of Times Receiver Alignment Was Used"
  )
```

### Motion ---------------------------------------

#### Proportion of Plays Where A WR Was Motioned ----------------

##### Entire League 
```{r}
motion <- plays_of_route_runners |>
  mutate(motioned = ifelse(!is.na(inMotionAtBallSnap), "Yes", "No")) |>  # NAs represent players that never motioned at all
  filter(motioned == "Yes")

motion_plays <- motion|>
  distinct(gameId, playId) |> 
  count() |> 
  rename(num_motion_plays = n)

total_plays <- plays_of_route_runners |> 
  distinct(gameId, playId) |> 
  count() |> 
  rename(num_total_plays = n)

calc_prop_motion <- bind_cols(motion_plays, total_plays) |> 
  mutate(motioned = num_motion_plays / num_total_plays,
         no_motion = 1 - motioned)

prop_motion <- calc_prop_motion |> 
  select(motioned, no_motion) |> 
  pivot_longer(motioned:no_motion, names_to = "motion", values_to = "prop") |> 
  mutate(motion = ifelse(motion == "motioned", "Yes", "No"))

```

```{r}
# Graph showing distribution of plays with motion
prop_motion |>   
  ggplot(aes(x = motion, y= prop, fill = motion)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Plays With Motion Or Not",
    x = "At Least One Player Motioned",
    y = "Proportion of Plays"
  )
```

##### Teams 
```{r}
team_motion <- motion |> 
  distinct(possessionTeam, gameId, playId) |> 
  count(possessionTeam) |> 
  rename(num_motion_plays = n)

team_total_plays <- plays_of_route_runners |> 
  distinct(possessionTeam, gameId, playId) |> 
  count(possessionTeam) |> 
  rename(num_total_plays = n)

prop_motion_team <- full_join(team_motion, team_total_plays)|> 
  mutate(prop_motioned = num_motion_plays / num_total_plays) 
```

```{r}
# Graph showing distribution of teams motioning
prop_motion_team |> 
  ggplot(aes(x = fct_reorder(possessionTeam, prop_motioned), y= prop_motioned, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.6618385, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold", angle = 90)) +
  scale_fill_nfl() +
  labs(
    title = "Distribution of Teams Motioning",
    x = "Team",
    y = "Prop of Plays Team Motioned At Least One Player"
  )
```

##### Players

```{r}
player_motion <- motion |> 
  distinct(displayName, gameId, playId) |> 
  count(displayName) |> 
  rename(num_motion_plays = n)

player_total_plays <- plays_of_route_runners |> 
  distinct(displayName, gameId, playId) |> 
  count(displayName) |> 
  rename(num_total_plays = n)
```

```{r}
prop_motion_player <- full_join(player_motion, player_total_plays)|> 
  mutate(prop_motioned = num_motion_plays / num_total_plays) |> 
  select(displayName, prop_motioned) |> 
  mutate(prop_motioned = replace_na(prop_motioned, 0)) |> 
  arrange(prop_motioned) 
```


#### Motion from the line of scrimmage (excludes players that were not set at the line) -------------

##### Entire League 

```{r}
# plays of players where they were set at the line
set_at_line <- plays_of_route_runners |> 
  mutate(motionSinceLineset = ifelse(motionSinceLineset == TRUE, 1, 0)) |>
  drop_na(motionSinceLineset) # removes players that were not set at the line of scrimmage
```

```{r}
league_motion_line <- set_at_line |> 
  group_by(motionSinceLineset) |> 
  summarise(num_plays = n()) |> 
  ungroup() |> 
  mutate(total_plays = sum(num_plays),
         prop = num_plays / total_plays,
         motionSinceLineset = ifelse(motionSinceLineset == 1, "Yes", "No")) 

```

```{r}
# Graph showing distribution of players set on the line that motioning
league_motion_line |>   
  ggplot(aes(x = motionSinceLineset, y= prop, fill = motionSinceLineset)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Players Set On Line That Motioned",
    x = "Players On Line Motioned",
    y = "Proportion of Players"
  )
```


##### Teams
```{r}
# the proportion of the times each player motioned when they were set at the line
team_prop_motion_from_line <- set_at_line|> 
  group_by(possessionTeam) |> 
  summarise(prop_in_motion_from_line = round(mean(motionSinceLineset), 3)) |> 
  arrange(prop_in_motion_from_line)
```

```{r}
# Graph showing distribution of teams motioning players set at the line of scrimmage 

team_prop_motion_from_line |> 
  ggplot(aes(x = fct_reorder(possessionTeam, prop_in_motion_from_line), y= prop_in_motion_from_line, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.1198928, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  scale_fill_nfl() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold", angle = 90)) +
  labs(
    title = "Distribution of Teams Motioning From Line",
    x = "Team",
    y = "Prop of Times Team Motioned Players At Line"
  )
```

##### Players
```{r}
# the proportion of the times each player motioned when they were set at the line
player_prop_motion_from_line <- set_at_line|> 
  group_by(displayName) |> 
  summarise(prop_in_motion_from_line = round(mean(motionSinceLineset), 3)) |> 
  arrange(prop_in_motion_from_line)

player_prop_motion_from_line
```


#### Motion and stay in motion at snap (excludes players that were not in motion prior to snap) --------------

##### Entire League
```{r}
# plays where a WR was in motion at some point
motion <- plays_of_route_runners |> 
  mutate(inMotionAtBallSnap = ifelse(inMotionAtBallSnap == TRUE, 1, 0)) |> 
  drop_na(inMotionAtBallSnap) # NAs represent players that never motioned at all
```

```{r}
league_motion_snap <- motion |> 
  group_by(inMotionAtBallSnap) |> 
  summarise(num_plays = n()) |> 
  ungroup() |> 
  mutate(total_plays = sum(num_plays),
         prop = num_plays / total_plays,
         inMotionAtBallSnap = ifelse(inMotionAtBallSnap == 1, "Yes", "No")) 
```

```{r}
# Graph showing distribution of plays with players on line motioning
league_motion_snap |>   
  ggplot(aes(x = inMotionAtBallSnap, y= prop, fill = inMotionAtBallSnap)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Players That Stayed In Motion At Snap",
    x = "Players Stayed In Motion At Snap?",
    y = "Proportion of Plays"
  )
```

##### Teams
```{r}
# table of proportion of the time that each team motions and keeps player in motion at snap
team_prop_motion <- motion|> 
  group_by(possessionTeam) |> 
  summarise(propInMotion = round(mean(inMotionAtBallSnap), 3)) |> 
  arrange(propInMotion)

```

```{r, fig.cap="The distribution is relatively even. MIA motions A LOT though. PHI and NE seem to motion less.}
# Visualization of proportion of the time that each team motions and keeps player in motion at snap - fix this so each team is represented by their team colors

team_prop_motion |> 
  ggplot(aes(x = fct_reorder(possessionTeam, propInMotion), y = propInMotion, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.1096526, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  scale_fill_nfl() +
  theme(axis.text.x = element_text(size = 8.5, angle = 90), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Proportion of Players Each NFL Team Motions & Stays In Motion At Snap ",
    x = "NFL Team",
    y = "Prop of Players That Stayed In Motion At Snap"
  )
```

##### Players
```{r}
# the proportion of the time each player is still in motion at snap when they previously motioned
player_prop_still_motion_at_snap <- motion|> 
  group_by(displayName) |> 
  summarise(prop_still_motion_at_snap = round(mean(inMotionAtBallSnap), 3)) |> 
  arrange(prop_still_motion_at_snap)

player_prop_still_motion_at_snap
```

#### Still In Motion At Snap At Line ----------------------------

##### Entire League

```{r}
line_snap_motion_league <- set_at_line |> 
  filter(motionSinceLineset == 1) |> 
  group_by(inMotionAtBallSnap) |> 
  summarise(num_players = n()) |> 
  mutate(total_players = sum(num_players),
         prop = num_players / total_players)
```

```{r}
# Graph showing distribution of players that were in motion from the line and stayed in motion at snap
line_snap_motion_league |>   
  ggplot(aes(x = inMotionAtBallSnap, y= prop, fill = inMotionAtBallSnap)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Players Set On Line That Motioned & Stayed In Motion At Snap",
    x = "Players On Line Stayed In Motion At Snap",
    y = "Proportion of Players"
  )
```

# Appendix ----------------------------------------------------------------------------

# Identifying which player position to focus on in our analysis --------------------------

```{r}
# identify all players (by nflID) that play WR 
WR <- players |> 
  filter(position == "WR") |> 
  distinct(nflId)

# identify all players that ran at least one route 
route_runners <- player_play |> 
  drop_na(routeRan) |> 
  distinct(nflId)

# identify which players were not WR that ran a route 
Non_WR_route_runners <- anti_join(route_runners, WR)
  
# identify which positions ran routes that were not WR and how often - notice there are QB, FB, TE, and RB as well
left_join(Non_WR_route_runners, players) |> 
  count(position)

# find out how many players play each position
players |> 
  count(position)
```

# Exploring the NAs left ------------------------------- 

```{r}
# See how many plays now have NAs for routes ran - 15,393
plays_of_route_runners |> 
  filter(routeRan == "NO ROUTE")

# There are no plays that were blown dead because of penalty
plays_of_route_runners |> 
  count(playNullifiedByPenalty)
```

## Figuring out when they might be blocking

```{r}
# All plays where the WR blocked a player- on 11 of them the WR still ran a route
plays_of_route_runners |> 
  filter(!is.na(blockedPlayerNFLId1))
```

```{r}
# Find out which routes have at least one WR that ran that route and one WR that did not run a route - makes sense that screen is one of the highest 
plays_of_route_runners |> 
  inner_join(no_route_ran) |> 
  filter(routeRan != "NO ROUTE") |> 
  count(routeRan)
```

## Identifying receivers the did not run routes when screen route was run
```{r}
screen_plays <- plays_of_route_runners |> 
  filter(routeRan == "SCREEN") |> 
  distinct(gameId, playId)

plays_of_route_runners |> 
  inner_join(screen_plays)|> 
  filter(routeRan == "NO ROUTE")|> 
  group_by(gameId, playId) |> 
  summarise(num_receivers_no_route = n()) |> 
  ungroup() |> 
  count(num_receivers_no_route)
```

## Figuring out if receivers are running routes on sacks
```{r}
plays_of_route_runners |> 
  filter(passResult == "S") |> 
  count(routeRan)
```

# Exploring Formations ------------------------------

## Jumbo 
```{r}

# filtering for plays where the formation was Jumbo
jumbo <- plays_of_route_runners |> 
  filter(offenseFormation == "JUMBO"
         ) |> 
  distinct(gameId, playId)

jumbo

# answered questions like - (1) are plays with off formation jumbo always runs (no) (2) how often in the jumbo formation do receivers run routes (3) how often do receivers block on jumbo plays
plays_of_route_runners |> 
  inner_join(jumbo) |> 
  filter(!is.na(passLength)) |> 
  #filter(routeRan != "NO ROUTE") |> 
  #filter(is.na(blockedPlayerNFLId1)) |> 
  #count(gameId, playId) |> 
  #count(n)

```

## Wildcat
```{r}

# filtering for plays where the formation was wildcat
wildcat <- plays_of_route_runners |> 
  filter(offenseFormation == "WILDCAT"
         ) |> 
  distinct(gameId, playId)

wildcat

# how many WR do not run routes when in the wildcat formation - can not remove wildcats because receivers are still running routes
plays_of_route_runners |> 
  inner_join(wildcat) |> 
  filter(routeRan != "NO ROUTE") |> 
  count(gameId, playId) |> 
  count(n) |> 
  rename(num_receivers = n,
         num_occurrences = nn)

```

# Figuring Out Differences / Relationships Between `motionSinceLineset` & `inMotionAtBallSnap` ------------

```{r}
plays_of_route_runners |> 
  count(inMotionAtBallSnap)



plays_of_route_runners |> 
  filter(motionSinceLineset == FALSE | is.na(motionSinceLineset),
        inMotionAtBallSnap == TRUE)

plays_of_route_runners |> 
  count(shiftSinceLineset)

```