---
title: "Route Prediction"
author: "Belle Schmidt"
date: "2024-11-19"
output: html_document
---

# Set Up ---------------------

```{r}
# reading in the packages necessary
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
```

```{r}
# gives me more information about the nflreadr package
library(help = "nflreadr")
```

```{r}
# reading in dataset
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
plays_of_WR <- read.csv("wr_plays.csv")
offense_prediction_plays <- read.csv("offense_prediction_plays.csv")

View(plays_of_WR)
```

```{r}
# reading in week by week tracking data
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_9.csv")
```

# Exploratory Data Analysis --------------------------------------------------------

## Distribution of Variables ------------------------------------

### Routes Ran -----------------


```{r}
# The number of times and the proportion of the time each route is run by WR (Table)
prop_route <- plays_of_WR |> 
  group_by(routeRan) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```


```{r, fig.cap= "There is a very large disproportion of the proportion of times each route is ran. Specifically a lot of the time receivers ran no route."}

# Plot of distribution of routes ran
prop_route |> 
  ggplot(aes(x = fct_reorder(routeRan, prop), y = prop, fill = routeRan)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Routes Ran By Wide Recievers",
    x = "Route Ran",
    y = "Proportion Of The Time Route Was Ran"
  )
```

#### Defining the Different Routes 
**Wheel:** When the reciever runs kind of an L shaped path. Out and then down field - most commonly run by RB

**Angle:** when the player cuts out 45 degrees and then takes a 90 degree turn and cuts back in (often RB, FB, or TE) - makes sense why few WR run it

**Screen:** Various types. More of a basic screen is where RB catches ball and lineman block. Then Jailbreak screen is where the receiver out wide is the target. Two receivers and a lineman are blockers. Bubble screen is where WR drops back slightly and two receivers block. Drop back screen looks like drop back pass but one WR runs route and other WR block. Could be TE too? 

**Flat:** when the receiver runs a quick short route straight then towards the sideline

**Corner:** when the receiver runs down the field and then turns outward 45degrees (opposite direction of post)

**Slant:** when the receiver runs shortly straight and then turns inward 45 degrees (like short version of post)

**Post:** when the receiver runs 10 to 20 yards down field and then turns 45 degrees and runs towards the goal posts 

**Out:** when the receiver runs down the field and then cuts out and runs parallel to the line of scrimmage

**In:** when the receiver runs down the field and then cuts in and runs parallel to the line of scrimmage

**Cross:** when the receiver crosses across the field

**Hitch:** when the receiver runs a set number of yards (normally 5 to 6) and then turns to face the QB 

**Go:** When the wide receiver runs straight down field as fast as they can

### Formations ------------------------

```{r}
# Table of the number of times and the proportion of the time each formation is used
prop_formation <- plays_of_WR |> 
  group_by(offenseFormation) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```


```{r}
# Graph showing distribution of formations 
prop_formation |> 
  ggplot(aes(x = fct_reorder(offenseFormation, prop), y = prop, fill = offenseFormation)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Formation Usage",
    x = "Formation",
    y = "Proportion of Time Formation Was Used"
  )
```

#### Defining the Different Formations

**Shotgun:** When the quarterback is several yards back from the center and has more time to read the defense. In this case, when the ball is snapped, the quarterback is not grabbing it directly from the center.

**Single back:** When the quarterback is under center (grabs the ball directly from the center) and there is a single RB about 5 yards back

**Empty:** When there are no backs in the backfield so the quarterback is the only player in the backfield

**I-Formation**: When the quarterback is under center. The fullback is directly behind him and the running back is behind him. It is typically a run play but can also be used for big gains because the wide receivers can get behind the defense faster. 

**Pistol:** When the quarterback is a couple yards behind the center (closer than for shotgun though) and there is a running back right behind him

**Wildcat:** when the ball is snapped to a position player other than the QB... might want to remove these

**Jumbo:** when an offensive lineup is designed for maximum blocking power (often close to goal line). Normally sub a receiver for an extra TE.


### Receiver Alignment ---------------------------

```{r}
# Table showing he number of times and the proportion of the time each receiver alignment is used
receiver_align_prop <- plays_of_WR |> 
  group_by(receiverAlignment) |> 
  summarise(occurances = n()) |> 
  mutate(prop = occurances / sum(occurances)) 
```

```{r}
# Graph showing distribution of receiver alignment 

receiver_align_prop |> 
  ggplot(aes(x = fct_reorder(receiverAlignment, prop), y= prop, fill = receiverAlignment)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Receiver Alignment Usage",
    x = "Receiver Alignment",
    y = "Proportion of Times Receiver Alignment Was Used"
  )
```

### Motion ---------------------------------------

#### Proportion of Plays Where A WR Was Motioned ----------------

##### Entire League 
```{r}
motion <- plays_of_WR |>
  mutate(motioned = ifelse(!is.na(inMotionAtBallSnap), "Yes", "No")) |>  # NAs represent players that never motioned at all
  filter(motioned == "Yes")

motion_plays <- motion|>
  distinct(gameId, playId) |> 
  count() |> 
  rename(num_motion_plays = n)

total_plays <- plays_of_WR |> 
  distinct(gameId, playId) |> 
  count() |> 
  rename(num_total_plays = n)

calc_prop_motion <- bind_cols(motion_plays, total_plays) |> 
  mutate(motioned = num_motion_plays / num_total_plays,
         no_motion = 1 - motioned)

prop_motion <- calc_prop_motion |> 
  select(motioned, no_motion) |> 
  pivot_longer(motioned:no_motion, names_to = "motion", values_to = "prop") |> 
  mutate(motion = ifelse(motion == "motioned", "Yes", "No"))

```

```{r}
# Graph showing distribution of plays with motion
prop_motion |>   
  ggplot(aes(x = motion, y= prop, fill = motion)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Plays With A WR Motioned Or Not",
    x = "At Least One WR Motioned",
    y = "Proportion of Plays"
  )
```

##### Teams 
```{r}
team_motion <- motion |> 
  distinct(possessionTeam, gameId, playId) |> 
  count(possessionTeam) |> 
  rename(num_motion_plays = n)

team_total_plays <- plays_of_WR |> 
  distinct(possessionTeam, gameId, playId) |> 
  count(possessionTeam) |> 
  rename(num_total_plays = n)

prop_motion_team <- full_join(team_motion, team_total_plays)|> 
  mutate(prop_motioned = num_motion_plays / num_total_plays) 
```


```{r}
# Graph showing distribution of teams motioning
prop_motion_team |> 
  ggplot(aes(x = fct_reorder(possessionTeam, prop_motioned), y= prop_motioned, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.6618385, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold", angle = 90)) +
  scale_fill_nfl() +
  labs(
    title = "Distribution of Teams Motioning A WR",
    x = "Team",
    y = "Prop of Plays Team Motioned At Least One WR"
  )
```

##### Players

```{r}
player_motion <- motion |> 
  distinct(displayName, gameId, playId) |> 
  count(displayName) |> 
  rename(num_motion_plays = n)

player_total_plays <- plays_of_WR |> 
  distinct(displayName, gameId, playId) |> 
  count(displayName) |> 
  rename(num_total_plays = n)
```

```{r}
prop_motion_player <- full_join(player_motion, player_total_plays)|> 
  mutate(prop_motioned = num_motion_plays / num_total_plays) |> 
  select(displayName, prop_motioned) |> 
  mutate(prop_motioned = replace_na(prop_motioned, 0)) |> 
  arrange(prop_motioned) 
```


#### Motion from the line of scrimmage (excludes players that were not set at the line) -------------

##### Entire League 

```{r}
# plays of players where they were set at the line
set_at_line <- plays_of_WR |> 
  mutate(motionSinceLineset = ifelse(motionSinceLineset == TRUE, 1, 0)) |>
  drop_na(motionSinceLineset) # removes players that were not set at the line of scrimmage
```

```{r}
league_motion_line <- set_at_line |> 
  group_by(motionSinceLineset) |> 
  summarise(num_plays = n()) |> 
  ungroup() |> 
  mutate(total_plays = sum(num_plays),
         prop = num_plays / total_plays,
         motionSinceLineset = ifelse(motionSinceLineset == 1, "Yes", "No")) 

```

```{r}
# Graph showing distribution of players set on the line that motioning
league_motion_line |>   
  ggplot(aes(x = motionSinceLineset, y= prop, fill = motionSinceLineset)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Players Set On Line That Motioned",
    x = "Players On Line Motioned",
    y = "Proportion of Players"
  )
```


##### Teams
```{r}
# the proportion of the times each player motioned when they were set at the line
team_prop_motion_from_line <- set_at_line|> 
  group_by(possessionTeam) |> 
  summarise(prop_in_motion_from_line = round(mean(motionSinceLineset), 3)) |> 
  arrange(prop_in_motion_from_line)
```

```{r}
# Graph showing distribution of teams motioning players set at the line of scrimmage 

team_prop_motion_from_line |> 
  ggplot(aes(x = fct_reorder(possessionTeam, prop_in_motion_from_line), y= prop_in_motion_from_line, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.1198928, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  scale_fill_nfl() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 10), axis.text = element_text(face = "bold", angle = 90)) +
  labs(
    title = "Distribution of Teams Motioning From Line",
    x = "Team",
    y = "Prop of Times Team Motioned Players At Line"
  )
```

##### Players
```{r}
# the proportion of the times each player motioned when they were set at the line
player_prop_motion_from_line <- set_at_line|> 
  group_by(displayName) |> 
  summarise(prop_in_motion_from_line = round(mean(motionSinceLineset), 3)) |> 
  arrange(prop_in_motion_from_line)

player_prop_motion_from_line
```


#### Motion and stay in motion at snap (excludes players that were not in motion prior to snap) --------------

##### Entire League
```{r}
# plays where a WR was in motion at some point
motion <- plays_of_WR |> 
  mutate(inMotionAtBallSnap = ifelse(inMotionAtBallSnap == TRUE, 1, 0)) |> 
  drop_na(inMotionAtBallSnap) # NAs represent players that never motioned at all
```

```{r}
league_motion_snap <- motion |> 
  group_by(inMotionAtBallSnap) |> 
  summarise(num_plays = n()) |> 
  ungroup() |> 
  mutate(total_plays = sum(num_plays),
         prop = num_plays / total_plays,
         inMotionAtBallSnap = ifelse(inMotionAtBallSnap == 1, "Yes", "No")) 
```

```{r}
# Graph showing distribution of plays with players on line motioning
league_motion_snap |>   
  ggplot(aes(x = inMotionAtBallSnap, y= prop, fill = inMotionAtBallSnap)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Distribution of Players That Stayed In Motion At Snap",
    x = "Players Stayed In Motion At Snap?",
    y = "Proportion of Plays"
  )
```

##### Teams
```{r}
# table of proportion of the time that each team motions and keeps player in motion at snap
team_prop_motion <- motion|> 
  group_by(possessionTeam) |> 
  summarise(propInMotion = round(mean(inMotionAtBallSnap), 3)) |> 
  arrange(propInMotion)

```

```{r, fig.cap="The distribution is relatively even. MIA motions A LOT though. PHI and NE seem to motion less.}
# Visualization of proportion of the time that each team motions and keeps player in motion at snap - fix this so each team is represented by their team colors

team_prop_motion |> 
  ggplot(aes(x = fct_reorder(possessionTeam, propInMotion), y = propInMotion, fill = possessionTeam)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.1096526, linetype = "dashed", color = "#FFC125", size = 1) +
  theme_bw() +
  scale_fill_nfl() +
  theme(axis.text.x = element_text(size = 8.5, angle = 90), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Proportion of Players Each NFL Team Motions & Stays In Motion At Snap ",
    x = "NFL Team",
    y = "Prop of Players That Stayed In Motion At Snap"
  )
```

##### Players
```{r}
# the proportion of the time each player is still in motion at snap when they previously motioned
player_prop_still_motion_at_snap <- motion|> 
  group_by(displayName) |> 
  summarise(prop_still_motion_at_snap = round(mean(inMotionAtBallSnap), 3)) |> 
  arrange(prop_still_motion_at_snap)

player_prop_still_motion_at_snap
```

#### Still In Motion At Snap At Line ----------------------------

##### Entire League

```{r}
line_snap_motion_league <- set_at_line |> 
  filter(motionSinceLineset == 1) |> 
  group_by(inMotionAtBallSnap) |> 
  summarise(num_players = n()) |> 
  mutate(total_players = sum(num_players),
         prop = num_players / total_players)
```

```{r}
# Graph showing distribution of players that were in motion from the line and stayed in motion at snap
line_snap_motion_league |>   
  ggplot(aes(x = inMotionAtBallSnap, y= prop, fill = inMotionAtBallSnap)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 8.5), legend.position = "none", plot.title = element_text(hjust = 0.5, face = "bold", size = 14), axis.title = element_text(face = "bold", size = 12), axis.text = element_text(face = "bold")) +
  labs(
    title = "Players Set On Line That Motioned & Stayed In Motion At Snap",
    x = "Players On Line Stayed In Motion At Snap",
    y = "Proportion of Players"
  )
```
