---
title: "BDB_Project"
author: "Neha Kotha"
date: "2024-12-29"
output: html_document
---

```{r}
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
library(ranger)
library(caret) 
```

```{r}
library(help = "nflreadr")
```

```{r}
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
tracking_week1 <- read.csv("tracking_week_1.csv")
tracking_week2 <- read.csv("tracking_week_2.csv")
tracking_week3 <- read.csv("tracking_week_3.csv")
tracking_week4 <- read.csv("tracking_week_4.csv")
tracking_week5 <- read.csv("tracking_week_5.csv")
tracking_week6 <- read.csv("tracking_week_6.csv")
tracking_week7 <- read.csv("tracking_week_7.csv")
tracking_week8 <- read.csv("tracking_week_8.csv")
tracking_week9 <- read.csv("tracking_week_9.csv")
```

# input information about a play to identify its specific playId. then, determine play based on game clock and play description 
```{r}
# identify specific play to analyze
specific_play <- plays |> 
  filter(possessionTeam == 'LA', 
         defensiveTeam == 'SF', 
         quarter == 2, down == 1, 
         yardsToGo == 10, preSnapHomeScore==7, preSnapVisitorScore==3) 
```

# PLAY ANIMATION
# filter data for one singular play: find nflId's of all players on the field and the ball data for play animation.
```{r}
# animation of MIA v CIN week 4, Q1 ----------------------------------
MIA_CIN_play <- tracking_week4 |> 
  filter(
    gameId == 2022092900,
    club %in% c("MIA", "CIN"),
    playId == 383
  ) |> 
  arrange(desc(s))

# all players on the field nflId's during pass play
players_on_field <- MIA_CIN_play |> 
  select(nflId, club, jerseyNumber) |> 
  distinct()

nfl_Ids <- players_on_field |> 
  pull(nflId)   

nfl_Ids

# filter tracking data to include specific players for this play and time range
MIA_CIN_play <- tracking_week4 |> 
  filter(
    gameId == 2022092900,  
    playId == 383, 
    nflId %in% nfl_Ids 
  )

MIA_CIN_play

MIA_CIN_play <- MIA_CIN_play|> 
  left_join(plays |> select(gameId, playId, absoluteYardlineNumber), 
            by = c("gameId", "playId"))
# filter for ball data
ball_data <- tracking_week4 |> 
  filter(
    gameId == 2022092900,
    playId == 383,
    is.na(nflId)  # row with NA nflId corresponds to the ball
  )

range(MIA_CIN_play$x)
range(MIA_CIN_play$y)
```

# animation of all players movement during the run play (1st Q, MIA vs. CIN, week 4)
```{r}
# define team colors
team_colors <- c("MIA" = "#00B5E2", "CIN" = "#FF5F00")

animation_all_players <- ggfootball(left_endzone = "#00B5E2", right_endzone = "#FF5F00", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = MIA_CIN_play, 
             aes(x = x, y = y, fill = club), size = 2, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add labels for jersey numbers
  geom_text(data = MIA_CIN_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # add the ball
  geom_point(data = ball_data, 
             aes(x = x, y = y), size = 1, shape = 23, fill = "brown", color = "black", stroke = 0.5, show.legend = FALSE) +
 
   # set theme and limits for the plot
  coord_fixed(xlim = c(55, 95), ylim = c(15, 55)) + 
  scale_color_manual(values = c("#FF5F00","#00B5E2")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#FF5F00","#00B5E2")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During MIA vs. CIN Pass Play",
       subtitle = "Frame: {frame_time}") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0),
    plot.subtitle = element_text(hjust = 0),
    axis.title = element_blank(),   
    axis.text = element_blank(),    
    axis.ticks = element_blank()
  )

# create the animation with transition by frameId
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

# display animation with proper frames per second 
total_frames <- length(unique(MIA_CIN_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)
```

##(after modeling is complete)
# play animation: color of each eligible receiver changes based on their most likely route ran at that frame of the play
```{r}
# Define a color palette for each route-- change later
route_colors <- c(
  "ANGLE" = "#FF0000", "CORNER" = "#FF7F00", "CROSS" = "#FFFF00", 
  "FLAT" = "#7FFF00", "GO" = "#00FF00", "HITCH" = "#00FFFF", 
  "IN" = "#007FFF", "OUT" = "#0000FF", "POST" = "#7F00FF", 
  "SCREEN" = "#FF00FF", "SLANT" = "#FF007F", "WHEEL" = "#BFBFBF", 
  "NO ROUTE" = "#000000"
)

# Generate predictions for each frame (mock example)
# Replace this with actual predictions from your model
test_MIA_CIN_play <- MIA_CIN_play |> 
  mutate(
    predicted_route = sample(c("ANGLE", "CORNER", "CROSS", "FLAT", "GO", "HITCH", "IN", "OUT", "POST", "SCREEN", "SLANT", "WHEEL", "NO ROUTE"), 
                             n(), replace = TRUE)
  )

# Assuming model is trained and stored in `prediction_model`
# Replace the placeholder code with actual predictions
#MIA_CIN_play <- MIA_CIN_play |> 
#  mutate(
#    predicted_route = predict(prediction_model, newdata = MIA_CIN_play)
#  )

animation_all_players <- ggfootball(left_endzone = "#00B5E2", right_endzone = "#FF5F00", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = test_MIA_CIN_play, 
             aes(x = x, y = y, fill = predicted_route), size = 2, shape = 21, color = "black", stroke = 0.5, show.legend = TRUE) + 
  # add labels for jersey numbers
  geom_text(data = test_MIA_CIN_play, 
            aes(x = x, y = y, label = jerseyNumber, color = "black"),
            size = 3, vjust = -1, show.legend = FALSE) +
  # add the ball
  geom_point(data = ball_data, 
             aes(x = x, y = y), size = 1, shape = 23, fill = "brown", color = "black", stroke = 0.5, show.legend = FALSE) +
  
  # Add dynamic fill scale for predicted routes
  scale_fill_manual(values = route_colors, name = "Predicted Route") +
  # set theme and limits for the plot
  coord_fixed(xlim = c(55, 95), ylim = c(15, 55)) + 
  labs(title = "Player Movements During MIA vs. CIN Pass Play",
       subtitle = "Frame: {frame_time}") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0),
    plot.subtitle = element_text(hjust = 0),
    axis.title = element_blank(),   
    axis.text = element_blank(),    
    axis.ticks = element_blank()
  )

# create the animation with transition by frameId
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

# display animation with proper frames per second 
total_frames <- length(unique(test_MIA_CIN_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)


```


# live line graph of player in motion's acceleration over time in play
```{r}

# Filter the data for this player
player_data <- MIA_CIN_play |> 
  filter(nflId == 48233) |> 
  arrange(frameId)  # Ensure time is in order

# Create an animated line graph for speed over time for the selected player
animation_speed_plot <- ggplot(player_data, aes(x = frameId, y = s)) +
  geom_line(color = "#00B5E2", linewidth = 0.5) +
  geom_point(color = "red", size = 1.5) +  # Highlight the current point
  labs(title = paste("Speed of Player", "displayName"), 
       x = "Time (Frame ID)", 
       y = "Speed (m/s)") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size = 12)
  ) +
  transition_reveal(frameId)  # Gradually reveal the line as frameId increases

animate(animation_speed_plot, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)

```

```{r}
# Total number of frames in the animation
total_frames <- length(unique(MIA_CIN_play$frameId))

# Animate the player movement (the field animation)
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)

# Animate the acceleration plot (line graph animation)
animate(acceleration_plot_animation, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)

```


# OLD CODE: includes full field view and arrows (orientation is incorrect)
# animation of all players movement during the run play (1st Q, MIA vs. CIN, week 4)
```{r}

# define team colors
team_colors <- c("MIA" = "#00B5E2", "CIN" = "#FF5F00")

animation_all_players <- ggfootball(left_endzone = "#00B5E2", right_endzone = "#FF5F00", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = MIA_CIN_play, 
             aes(x = x, y = y, fill = club), size = 2, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add labels for jersey numbers
  geom_text(data = MIA_CIN_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # add the ball
  geom_point(data = ball_data, 
             aes(x = x, y = y), size = 1, shape = 23, fill = "brown", color = "black", stroke = 0.5, show.legend = FALSE) +
  # add arrows for player orientation
  geom_segment(data = MIA_CIN_play, 
               aes(x = x, y = y, 
                   xend = pmin(pmax(x + cos(o * pi / 180) * 5, 0), 120),  # multiplier to adjust arrow length
                   yend = pmin(pmax(y + sin(o * pi / 180) * 5, 0), 53.3),   
                   color = club),
               arrow = arrow(type = "closed", length = unit(0.05, "inches")), size = 0.2, color = "black", show.legend = FALSE) +
  # set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  scale_color_manual(values = c("#FF5F00","#00B5E2")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#FF5F00","#00B5E2")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During Run Play", 
       subtitle = "Frame: {frame_time}", 
       x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

# create the animation with transition by frameId
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

# display animation with proper Frames Per Second (FPS)
total_frames <- length(unique(MIA_CIN_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)
```

# Get unique club names
```{r}
unique_clubs <- tracking_week1 |> 
  select(club) |> 
  distinct()
```
