---
title: "BDB_Project"
author: "Neha Kotha"
date: "2024-12-29"
output: html_document
---

```{r}
library(nflverse)
library(nflreadr)
library(nflplotR)
library(ggplot2)
library(tidyverse)
library(gganimate)
library(dplyr)
library(gganimate)
library(ggfootball)
library(ranger)
library(caret) 
library(help = "nflreadr")
```

```{r}
games <- read.csv("games.csv")
player_play <- read.csv("player_play.csv")
players <- read.csv("players.csv")
plays <- read.csv("plays.csv")
```

```{r}
tracking_w_1 <- read.csv("tracking_week_1.csv")
tracking_w_2 <- read.csv("tracking_week_2.csv")
tracking_w_3 <- read.csv("tracking_week_3.csv")
tracking_w_4 <- read.csv("tracking_week_4.csv")
tracking_w_5 <- read.csv("tracking_week_5.csv")
tracking_w_6 <- read.csv("tracking_week_6.csv")
tracking_w_7 <- read.csv("tracking_week_7.csv")
tracking_w_8 <- read.csv("tracking_week_8.csv")
tracking_w_9 <- read.csv("tracking_week_9.csv")
```

# identify specific playId. determine play based on game clock and play description 
```{r}
# identify specific play to analyze
specific_play <- plays |> 
  filter(possessionTeam == 'LA', 
         defensiveTeam == 'SF', 
         quarter == 2, down == 1, 
         yardsToGo == 10, preSnapHomeScore==7, preSnapVisitorScore==3) 
specific_play
```

# PLAY ANIMATION
# filter data for one singular play: find nflId's of all players on the field and the ball data for play animation.
```{r}
# animation of MIA v CIN week 4, Q1 ----------------------------------
MIA_CIN_play <- tracking_w_4 |> 
  filter(
    gameId == 2022092900,
    club %in% c("MIA", "CIN"),
    playId == 383
  ) |> 
  arrange(desc(s))

# all players on the field nflId's during pass play
players_on_field <- MIA_CIN_play |> 
  select(nflId, club, jerseyNumber) |> 
  distinct()

nfl_Ids <- players_on_field |> 
  pull(nflId)   

nfl_Ids

# filter tracking data to include specific players for this play and time range
MIA_CIN_play <- tracking_w_4 |> 
  filter(
    gameId == 2022092900,  
    playId == 383, 
    nflId %in% nfl_Ids 
  )

MIA_CIN_play

# filter for ball data
ball_data <- tracking_w_4 |> 
  filter(
    gameId == 2022092900,
    playId == 383,
    is.na(nflId)  # row with NA nflId corresponds to the ball
  )

range(MIA_CIN_play$x)
range(MIA_CIN_play$y)
```

# animation of all players movement during the run play (1st Q, MIA vs. CIN, week 4)
```{r}
# define team colors
team_colors <- c("MIA" = "#00B5E2", "CIN" = "#FF5F00")

animation_all_players <- ggfootball(left_endzone = "#00B5E2", right_endzone = "#FF5F00", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = MIA_CIN_play, 
             aes(x = x, y = y, fill = club), size = 2, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add labels for jersey numbers
  geom_text(data = MIA_CIN_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # add the ball
  geom_point(data = ball_data, 
             aes(x = x, y = y), size = 1, shape = 23, fill = "brown", color = "black", stroke = 0.5, show.legend = FALSE) +
 
   # set theme and limits for the plot
  coord_fixed(xlim = c(55, 95), ylim = c(15, 55)) + 
  scale_color_manual(values = c("#FF5F00","#00B5E2")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#FF5F00","#00B5E2")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During MIA vs. CIN Pass Play",
       subtitle = "Frame: {frame_time}") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0),
    plot.subtitle = element_text(hjust = 0),
    axis.title = element_blank(),   
    axis.text = element_blank(),    
    axis.ticks = element_blank()
  )

# create the animation with transition by frameId
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

# display animation with proper frames per second 
total_frames <- length(unique(MIA_CIN_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)
```

# OLD CODE: includes full field view and arrows (orientation is incorrect)
# animation of all players movement during the run play (1st Q, MIA vs. CIN, week 4)
```{r}

# define team colors
team_colors <- c("MIA" = "#00B5E2", "CIN" = "#FF5F00")

animation_all_players <- ggfootball(left_endzone = "#00B5E2", right_endzone = "#FF5F00", field_alpha = 0.8) +
  
  # plot all players positions over time as points
  geom_point(data = MIA_CIN_play, 
             aes(x = x, y = y, fill = club), size = 2, shape = 21, color = "black", stroke = 0.5, show.legend = FALSE) + 
  # add labels for jersey numbers
  geom_text(data = MIA_CIN_play, 
            aes(x = x, y = y, label = jerseyNumber, color = club),
            size = 3, vjust = -1, show.legend = FALSE) +
  # add the ball
  geom_point(data = ball_data, 
             aes(x = x, y = y), size = 1, shape = 23, fill = "brown", color = "black", stroke = 0.5, show.legend = FALSE) +
  # add arrows for player orientation
  geom_segment(data = MIA_CIN_play, 
               aes(x = x, y = y, 
                   xend = pmin(pmax(x + cos(o * pi / 180) * 5, 0), 120),  # multiplier to adjust arrow length
                   yend = pmin(pmax(y + sin(o * pi / 180) * 5, 0), 53.3),   
                   color = club),
               arrow = arrow(type = "closed", length = unit(0.05, "inches")), size = 0.2, color = "black", show.legend = FALSE) +
  # set theme and limits for the plot
  coord_fixed(xlim = c(0, 120), ylim = c(0, 53.3)) + 
  scale_color_manual(values = c("#FF5F00","#00B5E2")) +  # map team colors for the borders (arrows and text)
  scale_fill_manual(values = c("#FF5F00","#00B5E2")) +   # map team colors for the fill (inside of the points)
  labs(title = "Player Movements During Run Play", 
       subtitle = "Frame: {frame_time}", 
       x = "X Position (yards)", y = "Y Position (yards)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

# create the animation with transition by frameId
animation_all_players <- animation_all_players + 
  transition_time(frameId) +  # animate based on frameId
  ease_aes('linear') 

# display animation with proper Frames Per Second (FPS)
total_frames <- length(unique(MIA_CIN_play$frameId))
animate(animation_all_players, nframes = total_frames, fps = 25, duration = total_frames / 10,
        width = 800, height = 500, res = 150)
```

```{r}
# Get unique club names
unique_clubs <- tracking_w_1 |> 
  select(club) |> 
  distinct()
```

```{r}
```


```{r}
```



```{r}
```

```{r}
```

```{r}
```

